---
title: "Introduction to Random Walks in R"
description: "Learn the basics of random walks and their implementation in R"
order: 1
author: "Dogukan Nami Oztas"
date: "2025-05-16"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to Random Walks in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = 'center',
  fig.width = 8,
  fig.height = 6
)

# Load necessary packages with error handling
required_packages <- c("ggplot2", "dplyr", "gridExtra", "scales")

for (pkg in required_packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    stop(paste("Package", pkg, "is required but not installed. Please install it using: install.packages('", pkg, "')", sep = ""))
  }
}

library(ggplot2)
library(dplyr)
library(gridExtra)
library(scales)

# Source the functions from your R script
# Adjust the path if your Rmd file is in a different location relative to the R script
source("../R/01_random_walk_simulator.R")

# Set a consistent theme for all plots
theme_set(theme_minimal(base_size = 12))
```

## What is a Random Walk?

A **random walk** is a mathematical object, known as a stochastic or random process, that describes a path consisting of a succession of random steps. In our context, we'll focus on **one-dimensional random walks** where a particle (representing evidence) moves along a line, taking random steps at each time point.

### Mathematical Foundation

The discrete-time random walk model can be mathematically expressed as:

$$E_{t+1} = E_t + \mu + \epsilon_t$$

where: - $E_t$ is the evidence at time $t$ - $\mu$ is the **drift** (average step size per time unit) - $\epsilon_t \sim \mathcal{N}(0, \sigma^2)$ is random noise (Gaussian white noise)

The process continues until the evidence crosses one of two decision boundaries: - **Upper threshold**: $E_t \geq +a$ → Decision "A" (choice = 1) - **Lower threshold**: $E_t \leq -a$ → Decision "B" (choice = -1)

### Connection to Decision Making

Random walks are fundamental to understanding **decision-making models**, particularly the **Diffusion Decision Model (DDM)**. In psychological terms:

-   **Evidence accumulation**: Information gradually builds up over time
-   **Decision thresholds**: The amount of evidence needed to make a decision
-   **Drift**: Bias toward one option over another
-   **Noise**: Uncertainty and variability in the decision process

## Simulating a Single Random Walk Trial

Let's start by understanding how a single decision unfolds. We'll use the `simulate_random_walk_trial()` function to simulate individual trials and visualize the evidence accumulation process.

### Basic Single Trial Simulation

```{r single_rw_trials}
set.seed(123) # For reproducibility

# Example 1: Positive drift (bias toward upper threshold)
trial1 <- simulate_random_walk_trial(drift = 0.2, threshold = 5, sd_step = 1, return_path = TRUE)
cat("Trial 1 (positive drift):\n")
cat("Choice =", trial1$choice, ", RT (steps) =", trial1$rt, "\n\n")

# Example 2: Negative drift (bias toward lower threshold)
trial2 <- simulate_random_walk_trial(drift = -0.15, threshold = 5, sd_step = 1, return_path = TRUE)
cat("Trial 2 (negative drift):\n")
cat("Choice =", trial2$choice, ", RT (steps) =", trial2$rt, "\n\n")

# Example 3: No drift (unbiased)
trial3 <- simulate_random_walk_trial(drift = 0, threshold = 5, sd_step = 1, return_path = TRUE)
cat("Trial 3 (no drift):\n")
cat("Choice =", trial3$choice, ", RT (steps) =", trial3$rt, "\n")
```

### Visualizing Evidence Accumulation Paths

The power of the random walk model becomes clear when we visualize how evidence accumulates over time:

```{r visualize_paths, fig.height=8}
# Create plots for the three trials
plot_list <- list()

# Trial 1: Positive drift
plot_list[[1]] <- {
  df1 <- data.frame(
    time = 0:(length(trial1$evidence_path) - 1),
    evidence = trial1$evidence_path
  )
  
  ggplot(df1, aes(x = time, y = evidence)) +
    geom_line(color = "blue", size = 1) +
    geom_hline(yintercept = c(-5, 5), linetype = "dashed", color = "red", size = 1) +
    geom_hline(yintercept = 0, linetype = "dotted", color = "gray") +
    geom_point(data = df1[nrow(df1), ], color = "red", size = 3) +
    labs(title = "Trial 1: Positive Drift (μ = 0.2)",
         subtitle = paste("Choice:", trial1$choice, "| RT:", trial1$rt, "steps"),
         x = "Time Steps", y = "Evidence") +
    ylim(-6, 6) +
    annotate("text", x = max(df1$time) * 0.1, y = 5.5, label = "Upper Threshold", color = "red") +
    annotate("text", x = max(df1$time) * 0.1, y = -5.5, label = "Lower Threshold", color = "red")+
    xlim(-0.5, 15)
}

# Trial 2: Negative drift
plot_list[[2]] <- {
  df2 <- data.frame(
    time = 0:(length(trial2$evidence_path) - 1),
    evidence = trial2$evidence_path
  )
  
  ggplot(df2, aes(x = time, y = evidence)) +
    geom_line(color = "darkgreen", size = 1) +
    geom_hline(yintercept = c(-5, 5), linetype = "dashed", color = "red", size = 1) +
    geom_hline(yintercept = 0, linetype = "dotted", color = "gray") +
    geom_point(data = df2[nrow(df2), ], color = "red", size = 3) +
    labs(title = "Trial 2: Negative Drift (μ = -0.15)",
         subtitle = paste("Choice:", trial2$choice, "| RT:", trial2$rt, "steps"),
         x = "Time Steps", y = "Evidence") +
    ylim(-6, 6)+
    xlim(-0.5, 15)
}

# Trial 3: No drift
plot_list[[3]] <- {
  df3 <- data.frame(
    time = 0:(length(trial3$evidence_path) - 1),
    evidence = trial3$evidence_path
  )
  
  ggplot(df3, aes(x = time, y = evidence)) +
    geom_line(color = "purple", size = 1) +
    geom_hline(yintercept = c(-5, 5), linetype = "dashed", color = "red", size = 1) +
    geom_hline(yintercept = 0, linetype = "dotted", color = "gray") +
    geom_point(data = df3[nrow(df3), ], color = "red", size = 3) +
    labs(title = "Trial 3: No Drift (μ = 0)",
         subtitle = paste("Choice:", trial3$choice, "| RT:", trial3$rt, "steps"),
         x = "Time Steps", y = "Evidence") +
    ylim(-6, 6) +
    xlim(-0.5, 15)
}

# Arrange plots
do.call(grid.arrange, c(plot_list, ncol = 1))
```

**Key Observations:** - **Positive drift** tends to push evidence toward the upper threshold - **Negative drift** tends to push evidence toward the lower threshold\
- **No drift** creates a symmetric random walk where either outcome is equally likely - The **red dot** shows where the threshold was crossed - **Noise** creates the zigzag pattern in the evidence accumulation

## Exploring Parameter Effects

Let's explore how different parameters affect the random walk behavior:

```{r parameter_exploration, fig.height=10}
set.seed(456)

# Generate multiple trials with different parameters
param_trials <- list(
  low_noise = simulate_random_walk_trial(drift = 0.1, threshold = 5, sd_step = 0.5, return_path = TRUE),
  high_noise = simulate_random_walk_trial(drift = 0.1, threshold = 5, sd_step = 2, return_path = TRUE),
  low_threshold = simulate_random_walk_trial(drift = 0.1, threshold = 2, sd_step = 1, return_path = TRUE),
  high_threshold = simulate_random_walk_trial(drift = 0.1, threshold = 8, sd_step = 1, return_path = TRUE)
)

# Create comparison plots
param_plots <- list()

for (i in 1:length(param_trials)) {
  trial_name <- names(param_trials)[i]
  trial_data <- param_trials[[i]]
  
  df_temp <- data.frame(
    time = 0:(length(trial_data$evidence_path) - 1),
    evidence = trial_data$evidence_path
  )
  
  # Determine threshold value for plotting
  threshold_val <- switch(trial_name,
                         "low_threshold" = 2,
                         "high_threshold" = 8,
                         5)  # default
  
  param_plots[[i]] <- ggplot(df_temp, aes(x = time, y = evidence)) +
    geom_line(color = "steelblue", size = 1) +
    geom_hline(yintercept = c(-threshold_val, threshold_val), 
               linetype = "dashed", color = "red", size = 1) +
    geom_hline(yintercept = 0, linetype = "dotted", color = "gray") +
    geom_point(data = df_temp[nrow(df_temp), ], color = "red", size = 3) +
    labs(title = switch(trial_name,
                       "low_noise" = "Low Noise (σ = 0.5)",
                       "high_noise" = "High Noise (σ = 2.0)", 
                       "low_threshold" = "Low Threshold (a = 2)",
                       "high_threshold" = "High Threshold (a = 8)"),
         subtitle = paste("Choice:", trial_data$choice, "| RT:", trial_data$rt, "steps"),
         x = "Time Steps", y = "Evidence") +
    ylim(-10, 10) +
    xlim(-0.5, 20)
}

do.call(grid.arrange, c(param_plots, ncol = 2))
```

## Simulating a Random Walk Experiment

To understand the statistical properties of the random walk model, we need to simulate many trials. Let's conduct a comprehensive experiment:

```{r rw_experiment}
set.seed(789)
n_sim_trials <- 1000

# Simulate experiment with moderate parameters
rw_data <- simulate_random_walk_experiment(
  n_trials = n_sim_trials,
  threshold = 10,
  drift = 0.05, # Slight positive drift
  sd_step = 1.5,
  show_progress = FALSE
)

# Display basic statistics
cat("Experiment Summary:\n")
cat("Total trials simulated:", nrow(rw_data), "\n")
cat("Trials with valid decisions:", sum(!is.na(rw_data$choice)), "\n")
cat("Proportion of valid decisions:", round(mean(!is.na(rw_data$choice)), 3), "\n\n")

# Display the first few rows
knitr::kable(head(rw_data), 
             caption = "First 6 trials of the simulated random walk experiment.")
```

### Analyzing Experiment Results

#### **1. Choice Proportions and Accuracy**

```{r visualize_choice_proportions, fig.height=5}
# Filter out NA choices and create summary
valid_choices <- rw_data %>% 
  filter(!is.na(choice)) %>%
  mutate(choice_label = factor(choice, 
                              levels = c(-1, 1), 
                              labels = c("Lower Threshold (-1)", "Upper Threshold (1)")))

if(nrow(valid_choices) > 0) {
  # Calculate choice proportions
  choice_summary <- valid_choices %>%
    count(choice_label) %>%
    mutate(proportion = n / sum(n))

  # Create summary table
  knitr::kable(choice_summary %>% 
               select(choice_label, n, proportion), 
               col.names = c("Choice Outcome", "Count (N)", "Proportion"),
               caption = "Summary of Choice Proportions", 
               digits = 3)

  # Create visualization
  p_choice <- ggplot(choice_summary, aes(x = choice_label, y = proportion, fill = choice_label)) +
    geom_bar(stat = "identity", color = "black", alpha = 0.8, width = 0.6) +
    geom_text(aes(label = percent(proportion, accuracy = 0.1)), 
              vjust = -0.3, size = 4, fontface = "bold") +
    scale_y_continuous(labels = percent_format(accuracy = 1), 
                      limits = c(0, max(choice_summary$proportion) * 1.1)) +
    scale_fill_manual(values = c("tomato", "lightblue")) +
    labs(title = "Choice Proportions in Random Walk Experiment",
         subtitle = paste("n =", nrow(valid_choices), "trials with positive drift (μ = 0.05)"),
         x = "Decision Outcome", y = "Proportion of Trials") +
    theme(legend.position = "none")
  
  print(p_choice)
  
  # Statistical test for bias
  if (length(unique(valid_choices$choice)) == 2) {
    bias_test <- binom.test(sum(valid_choices$choice == 1), nrow(valid_choices), p = 0.5)
    cat("\nStatistical Test for Bias:\n")
    cat("Binomial test p-value:", round(bias_test$p.value, 6), "\n")
    cat("95% CI for upper threshold proportion:", 
        round(bias_test$conf.int[1], 3), "-", round(bias_test$conf.int[2], 3), "\n")
  }
} else {
  cat("No valid choices to analyze (all trials may have resulted in timeouts).\n")
}
```

With a positive drift (μ = 0.05), we expect more choices for the upper threshold, which represents the "correct" or "preferred" response in a decision-making context.

#### **2. Reaction Time (RT) Distributions**

```{r visualize_rt_distributions, fig.height=8}
if(nrow(valid_choices) > 0) {
  # RT summary statistics
  rt_summary <- valid_choices %>%
    group_by(choice_label) %>%
    summarise(
      N = n(),
      Mean_RT = round(mean(rt, na.rm = TRUE), 1),
      Median_RT = round(median(rt, na.rm = TRUE), 1),
      SD_RT = round(sd(rt, na.rm = TRUE), 1),
      Min_RT = min(rt, na.rm = TRUE),
      Max_RT = max(rt, na.rm = TRUE),
      .groups = "drop"
    )
  
  knitr::kable(rt_summary, 
               caption = "Summary Statistics for RTs by Choice Outcome")
  
  # Create RT distribution plots
  p1 <- ggplot(valid_choices, aes(x = rt, fill = choice_label)) +
    geom_histogram(bins = 30, alpha = 0.7, color = "black") +
    facet_wrap(~choice_label, scales = "free_y", ncol = 1) +
    scale_fill_manual(values = c("tomato", "lightblue")) +
    labs(title = "RT Distributions by Choice Outcome",
         x = "Reaction Time (Steps)", y = "Frequency",
         fill = "Choice") +
    theme(legend.position = "none")
  
  # Box plot comparison
  p2 <- ggplot(valid_choices, aes(x = choice_label, y = rt, fill = choice_label)) +
    geom_boxplot(alpha = 0.7, width = 0.6) +
    geom_jitter(width = 0.2, alpha = 0.3, size = 0.5) +
    scale_fill_manual(values = c("tomato", "lightblue")) +
    labs(title = "RT Comparison Between Choices",
         x = "Choice Outcome", y = "Reaction Time (Steps)") +
    theme(legend.position = "none")
  
  # Density overlay
  p3 <- ggplot(valid_choices, aes(x = rt, color = choice_label, fill = choice_label)) +
    geom_density(alpha = 0.3, size = 1) +
    scale_color_manual(values = c("darkred", "darkblue")) +
    scale_fill_manual(values = c("tomato", "lightblue")) +
    labs(title = "RT Density Distributions",
         x = "Reaction Time (Steps)", y = "Density",
         color = "Choice", fill = "Choice") +
    theme(legend.position = "bottom")
  
  # Arrange plots
  grid.arrange(p1, arrangeGrob(p2, p3, ncol = 2), ncol = 1, heights = c(2, 1))
  
  # Statistical comparison
  if (length(unique(valid_choices$choice)) == 2) {
    rt_test <- t.test(rt ~ choice, data = valid_choices)
    cat("\nRT Comparison Between Choices:\n")
    cat("t-test p-value:", round(rt_test$p.value, 6), "\n")
    cat("Mean difference (Upper - Lower):", round(rt_test$estimate[1] - rt_test$estimate[2], 2), "steps\n")
  }
}
```

**Key Insights from RT Analysis:** - RT distributions are typically **positively skewed** (long right tail) - **Faster RTs** often occur for the choice favored by drift - **Slower RTs** may indicate more difficult decisions or unfavorable drift - The **variability** in RTs reflects the stochastic nature of the process

#### **3. Speed-Accuracy Trade-off**

```{r speed_accuracy_analysis, fig.height=5}
if(nrow(valid_choices) > 0) {
  # Define "correct" responses as upper threshold (since drift is positive)
  accuracy_data <- valid_choices %>%
    mutate(
      correct = choice == 1,  # Upper threshold is "correct" with positive drift
      rt_bin = cut(rt, breaks = quantile(rt, probs = seq(0, 1, 0.2)), 
                   include.lowest = TRUE, labels = c("Fastest", "Fast", "Medium", "Slow", "Slowest"))
    )
  
  # Calculate accuracy by RT bins
  speed_acc_summary <- accuracy_data %>%
    group_by(rt_bin) %>%
    summarise(
      N = n(),
      Accuracy = mean(correct),
      Mean_RT = mean(rt),
      .groups = "drop"
    ) %>%
    filter(!is.na(rt_bin))
  
  knitr::kable(speed_acc_summary, 
               caption = "Speed-Accuracy Analysis: Accuracy by RT Quintiles",
               digits = 3)
  
  # Visualize speed-accuracy trade-off
  p_speed_acc <- ggplot(speed_acc_summary, aes(x = Mean_RT, y = Accuracy)) +
    geom_point(aes(size = N), color = "steelblue", alpha = 0.8) +
    geom_smooth(method = "lm", se = TRUE, color = "red", linetype = "dashed") +
    scale_y_continuous(labels = percent_format(), limits = c(0, 1)) +
    labs(title = "Speed-Accuracy Trade-off",
         subtitle = "Relationship between reaction time and decision accuracy",
         x = "Mean Reaction Time (Steps)", 
         y = "Accuracy (Proportion Correct)",
         size = "N Trials") +
    theme(legend.position = "bottom")
  
  print(p_speed_acc)
}
```

## Parameter Sensitivity Analysis

Let's explore how different parameter values affect model behavior:

```{r parameter_sensitivity, fig.height=10}
set.seed(999)

# Define parameter combinations to test
param_combinations <- expand.grid(
  drift = c(-0.1, 0, 0.1, 0.2),
  threshold = c(5, 10),
  sd_step = c(0.5, 1.5)
)

# Simulate smaller experiments for each combination
sensitivity_results <- list()

for (i in 1:nrow(param_combinations)) {
  params <- param_combinations[i, ]
  
  sim_data <- simulate_random_walk_experiment(
    n_trials = 200,  # Smaller n for speed
    drift = params$drift,
    threshold = params$threshold,
    sd_step = params$sd_step,
    show_progress = FALSE
  )
  
  # Calculate summary statistics
  valid_trials <- sim_data[!is.na(sim_data$choice), ]
  
  if (nrow(valid_trials) > 0) {
    sensitivity_results[[i]] <- data.frame(
      drift = params$drift,
      threshold = params$threshold,
      sd_step = params$sd_step,
      prop_upper = mean(valid_trials$choice == 1),
      mean_rt = mean(valid_trials$rt),
      prop_valid = nrow(valid_trials) / 200
    )
  }
}

# Combine results
sensitivity_df <- do.call(rbind, sensitivity_results)

# Create visualizations
if (nrow(sensitivity_df) > 0) {
  # Choice proportion heatmap
  p1 <- ggplot(sensitivity_df, aes(x = factor(drift), y = factor(threshold), fill = prop_upper)) +
    geom_tile(color = "white") +
    facet_wrap(~paste("Noise σ =", sd_step)) +
    scale_fill_gradient2(low = "red", mid = "white", high = "blue", midpoint = 0.5,
                        labels = percent_format()) +
    labs(title = "Choice Bias (Proportion Choosing Upper Threshold)",
         x = "Drift (μ)", y = "Threshold (a)", fill = "Prop. Upper") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  # Mean RT heatmap  
  p2 <- ggplot(sensitivity_df, aes(x = factor(drift), y = factor(threshold), fill = mean_rt)) +
    geom_tile(color = "white") +
    facet_wrap(~paste("Noise σ =", sd_step)) +
    scale_fill_gradient(low = "yellow", high = "darkred") +
    labs(title = "Mean Reaction Time",
         x = "Drift (μ)", y = "Threshold (a)", fill = "Mean RT") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  grid.arrange(p1, p2, ncol = 1)
}
```

## Summary and Next Steps

This vignette introduced the fundamental concepts of random walks and their implementation in R. We've covered:

### **Key Concepts Learned:**

1.  **Mathematical foundation** of discrete-time random walks
2.  **Parameter effects**: drift, threshold, and noise
3.  **Simulation methods** for single trials and experiments\
4.  **Statistical analysis** of choice proportions and RT distributions
5.  **Speed-accuracy trade-offs** in decision making
6.  **Parameter sensitivity** analysis

### **Important Insights:**

-   **Drift** creates bias toward one choice option
-   **Threshold** controls the speed-accuracy trade-off
-   **Noise** increases variability in both choices and RTs
-   **RT distributions** are typically positively skewed
-   The model exhibits realistic **speed-accuracy relationships**

### **Connection to Cognitive Science:**

Random walks provide the foundation for understanding: - **Evidence accumulation** in perceptual decisions - **Response time** distributions in choice tasks - **Confidence** and decision certainty - **Individual differences** in decision-making strategies
