---
title: "Simulating the Diffusion Decision Model (DDM)"
description: "How to simulate the DDM in R and understand its parameters."
order: 2
author: "Dogukan Nami Oztas"
date: "2025-05-16"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulating the Diffusion Decision Model (DDM)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = 'center',
  fig.width = 8,
  fig.height = 6
)
# Load necessary packages
library(ggplot2)
library(dplyr) # For data manipulation
library(tidyr) # For data wrangling if needed
library(gridExtra) # For arranging plots

# Source the functions from your DDM R script
source("../R/02_ddm_simulator_basic.R")
source("../R/utils/plot_basic_ddm_path_rt.R")
source("../R/utils/plot_parameter_effects.R")
```

## Introduction to the Diffusion Decision Model (DDM)

The Diffusion Decision Model (DDM) is a prominent mathematical model used in psychology and cognitive neuroscience to account for reaction times (RTs) and accuracy in two-alternative forced-choice (2AFC) decision tasks. It conceptualizes decision-making as a process of noisy evidence accumulation over time towards one of two decision boundaries.

The DDM is essentially a continuous-time version of the random walk model we explored in the previous vignette. Instead of discrete steps, evidence is assumed to accumulate continuously, influenced by both a systematic "drift" (reflecting the quality of stimulus information) and random "noise" (reflecting moment-to-moment variability in processing).

### Why Use the DDM?

The DDM is particularly valuable because it:

1.  **Jointly explains accuracy and RT**: Unlike models that treat these separately, the DDM provides a unified account
2.  **Captures individual differences**: Parameters can vary across participants and conditions
3.  **Has strong theoretical foundation**: Based on optimal decision-making principles
4.  **Provides mechanistic insights**: Links behavioral data to underlying cognitive processes

### Mathematical Foundation

The DDM is mathematically described by a stochastic differential equation:

$$dx(t) = v \cdot dt + s \cdot dW(t)$$

Where: - $x(t)$ is the evidence state at time $t$ - $v$ is the drift rate (systematic accumulation) - $s$ is the diffusion coefficient (noise) - $dW(t)$ is a Wiener process increment

This vignette will demonstrate how to simulate the DDM using the functions created in `R/02_ddm_simulator_basic.R`.

## Key DDM Parameters

Our DDM simulator, `simulate_diffusion_trial()`, uses the Feller (1968) convention, where evidence accumulates between a lower boundary at 0 and an upper boundary at $a$.

```{r parameter_illustration, echo=FALSE, fig.height=4, fig.width=6}
# Create a simple illustration of DDM parameters
set.seed(123)
example_params <- list(v = 0.2, a = 1.0, z = 0.4, s = 0.3, dt = 0.01)

# Simulate a single path for illustration
illustration_trial <- simulate_diffusion_trial_with_path(
  v = example_params$v,
  a = example_params$a, 
  z = example_params$z,
  s = example_params$s,
  dt = example_params$dt
)

# Create illustration plot
ggplot(illustration_trial$path_data, aes(x = time_s, y = evidence)) +
  geom_line(color = "steelblue", linewidth = 1.2) +
  geom_hline(yintercept = 0, linetype = "solid", color = "red", linewidth = 1) +
  geom_hline(yintercept = example_params$a, linetype = "solid", color = "red", linewidth = 1) +
  geom_hline(yintercept = example_params$z, linetype = "dashed", color = "green", linewidth = 1) +
  geom_segment(aes(x = 0, y = example_params$z, 
                   xend = max(time_s), yend = example_params$z + example_params$v * max(time_s)),
               color = "orange", linewidth = 1, linetype = "dotted") +
  annotate("text", x = max(illustration_trial$path_data$time_s) * 0.9, 
           y = example_params$a + 0.05, label = "Upper boundary (a)", color = "red") +
  annotate("text", x = max(illustration_trial$path_data$time_s) * 0.9, 
           y = -0.05, label = "Lower boundary (0)", color = "red") +
  annotate("text", x = max(illustration_trial$path_data$time_s) * 0.1, 
           y = example_params$z + 0.05, label = "Starting point (z)", color = "green") +
  annotate("text", x = max(illustration_trial$path_data$time_s) * 0.5, 
           y = example_params$z + example_params$v * max(illustration_trial$path_data$time_s) * 0.5 + 0.1, 
           label = "Drift rate (v)", color = "orange") +
  labs(title = "DDM Parameters Illustrated",
       x = "Time (s)", y = "Evidence") +
  theme_minimal() +
  theme(legend.position = "none")
```

### Parameter Descriptions

-   **v (Drift Rate):**
    -   The average rate of evidence accumulation per unit time
    -   Positive $v$ indicates drift towards the upper boundary $a$
    -   Negative $v$ indicates drift towards the lower boundary 0
    -   Magnitude reflects stimulus discriminability: $|v| \propto$ signal strength
    -   **Typical range**: -2 to +2 (in standardized units)
-   **a (Threshold Separation):**
    -   Distance between decision boundaries (0 and $a$)
    -   Controls the speed-accuracy tradeoff
    -   Higher $a$ → slower, more accurate responses
    -   Lower $a$ → faster, less accurate responses
    -   **Typical range**: 0.5 to 2.5
-   **z (Starting Point):**
    -   Initial evidence state: $0 < z < a$
    -   $z = a/2$: unbiased starting point
    -   $z > a/2$: bias toward upper boundary
    -   $z < a/2$: bias toward lower boundary
    -   **Typical range**: $z/a$ from 0.3 to 0.7
-   **s (Noise/Diffusion Coefficient):**
    -   Standard deviation of moment-to-moment variability
    -   Often fixed to 0.1 for parameter scaling
    -   Higher $s$ increases RT variability
    -   **Typical value**: 0.1 (fixed)
-   **ter (Non-Decision Time):**
    -   Time for encoding and motor processes
    -   Added to decision time for total RT
    -   **Typical range**: 0.1 to 0.5 seconds
-   **dt (Time Step):**
    -   Simulation granularity for Euler-Maruyama approximation
    -   Smaller $dt$ → higher accuracy, longer computation
    -   **Recommended**: 0.001 to 0.01 seconds

## Simulating a Single DDM Trial

The `simulate_diffusion_trial()` function simulates one instance of the evidence accumulation process.

```{r single_ddm_trial_example}
set.seed(102) # For reproducibility

# Example 1: Positive drift, unbiased start
trial_A <- simulate_diffusion_trial(v = 0.2, a = 1.0, z = 0.5, s = 0.1, ter = 0.15)
cat("Trial A: Choice =", trial_A$choice, "(1=Upper, 0=Lower), RT =", round(trial_A$rt, 3), "s\n")

# Example 2: Negative drift, biased towards upper
trial_B <- simulate_diffusion_trial(v = -0.2, a = 1.0, z = 0.7, s = 0.1, ter = 0.15)
cat("Trial B: Choice =", trial_B$choice, ", RT =", round(trial_B$rt, 3), "s\n")

# Example 3: No drift, unbiased start
trial_C <- simulate_diffusion_trial(v = 0.0, a = 1.0, z = 0.5, s = 0.1, ter = 0.15)
cat("Trial C: Choice =", trial_C$choice, ", RT =", round(trial_C$rt, 3), "s\n")

# Example 4: High threshold (conservative)
trial_D <- simulate_diffusion_trial(v = 0.2, a = 1.8, z = 0.9, s = 0.1, ter = 0.15)
cat("Trial D: Choice =", trial_D$choice, ", RT =", round(trial_D$rt, 3), "s\n")
```

**Interpretation**: - Trial A: Positive drift should favor upper boundary (choice = 1) - Trial B: Despite negative drift, high starting point might still reach upper boundary - Trial C: With no systematic drift, choice depends purely on noise - Trial D: High threshold leads to longer, more accurate decisions

## Understanding Parameter Effects Through Simulation

Let's systematically explore how each parameter affects DDM behavior:

### Effect of Drift Rate (v)

```{r drift_rate_effects, fig.height=8}
set.seed(1001)
drift_rates <- c(-0.3, -0.1, 0, 0.1, 0.3)
n_trials_per_condition <- 500

# Simulate data for different drift rates
drift_results <- data.frame()
for(v in drift_rates) {
  sim_data <- simulate_diffusion_experiment(
    n_trials = n_trials_per_condition,
    v = v, a = 0.7, z = 0.35, s = 0.2, ter = 0.1
  )
  sim_data$drift_rate <- v
  drift_results <- rbind(drift_results, sim_data)
}

# Remove any NA trials
drift_results <- drift_results %>% filter(!is.na(choice))

# Plot 1: Choice proportions by drift rate
p1 <- drift_results %>%
  group_by(drift_rate) %>%
  summarise(prop_upper = mean(choice == 1), .groups = 'drop') %>%
  ggplot(aes(x = drift_rate, y = prop_upper)) +
  geom_point(size = 3, color = "steelblue") +
  geom_line(linewidth = 1, color = "steelblue") +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "red") +
  labs(title = "Choice Probability vs. Drift Rate",
       x = "Drift Rate (v)", y = "P(Upper Boundary)") +
  theme_minimal()

# Plot 2: Mean RT by drift rate and choice
p2 <- drift_results %>%
  group_by(drift_rate, choice) %>%
  summarise(mean_rt = mean(rt), .groups = 'drop') %>%
  ggplot(aes(x = drift_rate, y = mean_rt, color = factor(choice))) +
  geom_point(size = 3) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = c("0" = "salmon", "1" = "steelblue"),
                     labels = c("Lower", "Upper"),
                     name = "Boundary") +
  labs(title = "Mean RT vs. Drift Rate",
       x = "Drift Rate (v)", y = "Mean RT (s)") +
  theme_minimal()

# Combine plots
grid.arrange(p1, p2, ncol = 1)
```

**Key Observations**: - Higher drift rates increase accuracy (choice probability) - RTs are typically faster for choices in the direction of drift - The relationship between drift rate and choice follows a sigmoid (psychometric) function

### Effect of Threshold (a)

```{r threshold_effects}
set.seed(1002)
thresholds <- c(0.4, 0.6, 0.8, 1.0, 1.2, 1.4, 1.6)

threshold_results <- data.frame()
for(a in thresholds) {
  sim_data <- simulate_diffusion_experiment(
    n_trials = n_trials_per_condition,
    v = 0.15, a = a, z = a/2, s = 0.25, ter = 0.1  # Keep z proportional to a
  )
  sim_data$threshold <- a
  threshold_results <- rbind(threshold_results, sim_data)
}

threshold_results <- threshold_results %>% filter(!is.na(choice))

# Plot accuracy and mean RT vs threshold
p3 <- threshold_results %>%
  group_by(threshold) %>%
  summarise(accuracy = mean(choice == 1), .groups = 'drop') %>%  # Assuming v > 0, so upper is "correct"
  ggplot(aes(x = threshold, y = accuracy)) +
  geom_point(size = 3, color = "darkgreen") +
  geom_line(linewidth = 1, color = "darkgreen") +
  labs(title = "Accuracy vs. Threshold",
       x = "Threshold (a)", y = "Accuracy") +
  theme_minimal()

p4 <- threshold_results %>%
  group_by(threshold) %>%
  summarise(mean_rt = mean(rt), .groups = 'drop') %>%
  ggplot(aes(x = threshold, y = mean_rt)) +
  geom_point(size = 3, color = "purple") +
  geom_line(linewidth = 1, color = "purple") +
  labs(title = "Mean RT vs. Threshold",
       x = "Threshold (a)", y = "Mean RT (s)") +
  theme_minimal()

grid.arrange(p3, p4, ncol = 2)
```

**Key Observations**: - Higher thresholds lead to higher accuracy but longer RTs - This demonstrates the fundamental speed-accuracy tradeoff in decision-making

### Using Utility Functions for Parameter Exploration

Our enhanced DDM package includes utility functions that make parameter exploration more systematic and efficient:

```{r parameter_grid_example}
# Create a parameter grid for systematic exploration
param_grid <- create_parameter_grid(
  v_values = c(0.1, 0.2, 0.3),
  a_values = c(0.8, 1.0, 1.2),
  z_values = c(0.4, 0.5, 0.6),
  z_as_proportion = FALSE  # Use absolute z values
)

# Display first few parameter combinations
knitr::kable(head(param_grid), caption = "Example Parameter Grid for Systematic Exploration")

# Create organized parameter objects
easy_params <- create_ddm_params(v = 0.3, a = 1.0, z = 0.5, name = "Easy Condition")
hard_params <- create_ddm_params(v = 0.1, a = 1.0, z = 0.5, name = "Hard Condition")

print(easy_params)
print(hard_params)
```

## Simulating a DDM Experiment

To understand the model's predictions for RT distributions and choice probabilities, we simulate many trials using `simulate_diffusion_experiment()`.

```{r ddm_experiment_simulation}
set.seed(202)
n_sim_trials <- 2000 # Use a good number of trials for smooth distributions

# Define a set of parameters
params <- list(
  v = 0.1,      # Moderate positive drift
  a = 0.7,       # Threshold
  z = 0.35,       # Unbiased start (a/2)
  s = 0.2,       # Standard noise
  ter = 0.2,     # Non-decision time
  dt = 0.01     # Time step (higher precision)
)

ddm_data <- simulate_diffusion_experiment(
  n_trials = n_sim_trials,
  v = params$v,
  a = params$a,
  z = params$z,
  s = params$s,
  ter = params$ter,
  dt = params$dt
)

# Display the first few rows
knitr::kable(head(ddm_data, 10), digits = 3, 
             caption = "First 10 trials of the simulated DDM experiment.")

# Summary of the data using our utility function
summary_table <- create_summary_table(ddm_data, correct_response = 1)
knitr::kable(summary_table, caption = "Comprehensive Summary of DDM Simulation Results")
```

### Analyzing DDM Experiment Results

#### **1. Choice Proportions and Accuracy**

```{r choice_analysis, fig.height=4}
# Filter out NA choices (if max_decision_time was hit)
valid_ddm_choices <- na.omit(ddm_data$choice)

choice_counts_ddm <- table(valid_ddm_choices)
choice_proportions_ddm <- prop.table(choice_counts_ddm)

cat("Choice Counts:\n")
print(choice_counts_ddm)
cat("\nChoice Proportions:\n")
knitr::kable(as.data.frame(choice_proportions_ddm), 
             col.names = c("Choice", "Proportion"), 
             caption = "DDM Choice Proportions")

# Create a more informative bar plot
choice_df <- data.frame(
  Boundary = c("Lower (0)", "Upper (1)"),
  Proportion = as.numeric(choice_proportions_ddm),
  Count = as.numeric(choice_counts_ddm)
)

ggplot(choice_df, aes(x = Boundary, y = Proportion, fill = Boundary)) +
  geom_col(alpha = 0.8, color = "black") +
  geom_text(aes(label = paste0(round(Proportion*100, 1), "%\n(n=", Count, ")")), 
            vjust = -0.5, size = 4) +
  scale_fill_manual(values = c("salmon", "steelblue")) +
  labs(title = paste("Choice Proportions (v =", params$v, ")"),
       x = "Decision Boundary", y = "Proportion") +
  theme_minimal() +
  theme(legend.position = "none") +
  ylim(0, max(choice_df$Proportion) * 1.2)
```

**Expected Result**: With positive drift (v = `r params$v`), we expect more choices for the upper boundary (Choice 1), which we observe as `r round(choice_proportions_ddm[2]*100, 1)`% of trials.

#### **2. Reaction Time (RT) Distributions**

The DDM is well-known for predicting positively skewed RT distributions - a hallmark of decision-making data.

```{r ddm_rt_distributions, fig.height=8}
# Filter out NA RTs
valid_ddm_rt_data <- ddm_data %>% filter(!is.na(rt))

if(nrow(valid_ddm_rt_data) > 0) {
  
  # Create comprehensive RT analysis
  # Plot 1: Decision time distributions
  p_decision <- ggplot(valid_ddm_rt_data, aes(x = decision_time, fill = factor(choice))) +
    geom_histogram(binwidth = 0.02, alpha = 0.7, position = "identity", color = "black") +
    facet_wrap(~factor(choice, labels = c("Lower Boundary (0)", "Upper Boundary (1)"))) +
    scale_fill_manual(values = c("salmon", "steelblue")) +
    labs(title = "Decision Time Distributions",
         x = "Decision Time (s)", y = "Frequency", fill = "Choice") +
    theme_minimal() +
    theme(legend.position = "none")
  
  # Plot 2: Total RT distributions  
  p_total <- ggplot(valid_ddm_rt_data, aes(x = rt, fill = factor(choice))) +
    geom_histogram(binwidth = 0.02, alpha = 0.7, position = "identity", color = "black") +
    facet_wrap(~factor(choice, labels = c("Lower Boundary (0)", "Upper Boundary (1)"))) +
    scale_fill_manual(values = c("salmon", "steelblue")) +
    labs(title = "Total RT Distributions (Decision Time + Non-Decision Time)",
         x = "Total RT (s)", y = "Frequency", fill = "Choice") +
    theme_minimal() +
    theme(legend.position = "none")
  
  grid.arrange(p_decision, p_total, ncol = 1)

  # Summary statistics for RTs
  rt_summary_ddm <- valid_ddm_rt_data %>%
    group_by(choice) %>%
    summarise(
      N = n(),
      Mean_RT = mean(rt),
      Median_RT = median(rt),
      SD_RT = sd(rt),
      Min_RT = min(rt),
      Max_RT = max(rt),
      Skewness = (mean(rt) - median(rt)) / sd(rt),  # Simple skewness measure
      .groups = 'drop'
    )

  knitr::kable(rt_summary_ddm, digits = 3, 
               caption = "Summary Statistics for DDM RTs by Choice")
  
  # Highlight the skewness
  cat("\n=== Skewness Analysis ===\n")
  cat("Positive skew (Mean > Median) indicates right-skewed distributions\n")
  for(i in 1:nrow(rt_summary_ddm)) {
    choice_val <- rt_summary_ddm$choice[i]
    skew_val <- rt_summary_ddm$Skewness[i]
    cat("Choice", choice_val, ": Skewness =", round(skew_val, 3), 
        ifelse(skew_val > 0, "(right-skewed)", "(left-skewed)"), "\n")
  }

} else {
  cat("No valid trials with RTs to plot (all may have hit max_decision_time).\n")
}
```

**Key Observations**: - Both distributions show positive skew (right tail), characteristic of DDM predictions - "Correct" responses (upper boundary, given positive drift) tend to be faster on average - The addition of non-decision time shifts the entire distribution rightward

## Visualizing Evidence Accumulation Paths

To gain intuitive understanding of the DDM, let's visualize actual evidence accumulation paths.

### Individual Trial Paths

```{r ddm_path_simulation_setup, fig.height=6}
path_params <- list(
  v = 0.1,    # Moderate drift for interesting paths
  a = 1.0,    # Threshold
  z = 0.5,    # Unbiased start
  s = 0.3,    # Moderate noise for variability
  dt = 0.01,
  ter = 0.1   
)

n_paths_to_plot <- 8 # More paths for better illustration
all_paths_data <- vector("list", n_paths_to_plot)

set.seed(343) # For reproducibility
for (i in 1:n_paths_to_plot) {
  trial_with_path <- simulate_diffusion_trial_with_path(
    v = path_params$v,
    a = path_params$a,
    z = path_params$z,
    s = path_params$s,
    dt = path_params$dt,
    ter = path_params$ter
  )
  trial_with_path$path_data$trial_id <- paste("Trial", i)
  trial_with_path$path_data$final_choice <- trial_with_path$choice
  trial_with_path$path_data$final_rt <- trial_with_path$rt
  all_paths_data[[i]] <- trial_with_path$path_data
}

plot_df_paths <- bind_rows(all_paths_data)
threshold_a <- path_params$a

if(nrow(plot_df_paths) > 0) {
  ggplot(plot_df_paths, aes(x = time_s, y = evidence, group = trial_id, 
                           color = factor(final_choice))) +
    geom_line(alpha = 0.8, linewidth = 0.8) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "black", linewidth = 1) +
    geom_hline(yintercept = threshold_a, linetype = "dashed", color = "black", linewidth = 1) +
    geom_hline(yintercept = path_params$z, linetype = "dotted", color = "grey40", linewidth = 0.8) +
    # Add drift line
    geom_segment(aes(x = 0, y = path_params$z, 
                     xend = max(time_s, na.rm = TRUE), 
                     yend = path_params$z + path_params$v * max(time_s, na.rm = TRUE)),
                 color = "orange", linewidth = 1.5, linetype = "longdash",
                 inherit.aes = FALSE) +
    annotate("text", x = max(plot_df_paths$time_s, na.rm = TRUE) * 0.95, 
             y = threshold_a + 0.05, label = paste("Upper boundary (a =", threshold_a, ")"), 
             hjust = 1, size = 3.5, fontface = "bold") +
    annotate("text", x = max(plot_df_paths$time_s, na.rm = TRUE) * 0.95, 
             y = -0.05, label = "Lower boundary (0)", 
             hjust = 1, size = 3.5, fontface = "bold") +
    annotate("text", x = 0.1, y = path_params$z + 0.08, 
             label = paste("Starting point (z =", path_params$z, ")"), 
             hjust = 0, size = 3.5, color = "grey40") +
    annotate("text", x = max(plot_df_paths$time_s, na.rm = TRUE) * 0.6, 
             y = path_params$z + path_params$v * max(plot_df_paths$time_s, na.rm = TRUE) * 0.6 + 0.1, 
             label = paste("Average drift (v =", path_params$v, ")"), 
             hjust = 0.5, size = 3.5, color = "orange", fontface = "bold") +
    scale_color_manual(values = c("0" = "salmon", "1" = "steelblue", "NA" = "grey"),
                       labels = c("0" = "Lower Boundary", "1" = "Upper Boundary", "NA"= "Timeout"),
                       name = "Trial Outcome") +
    labs(title = "DDM Evidence Accumulation Paths",
         subtitle = paste("Parameters: v =", path_params$v, ", a =", path_params$a, 
                         ", z =", path_params$z, ", s =", path_params$s),
         x = "Decision Time (s)", y = "Accumulated Evidence") +
    theme_minimal() +
    theme(legend.position = "top",
          plot.title = element_text(size = 14, face = "bold"),
          plot.subtitle = element_text(size = 12))
} else {
  cat("No path data to plot.\n")
}
```

### Comprehensive Path and RT Visualization

Our custom visualization function provides an integrated view of the decision process:

```{r advanced_path_visualization, fig.height=8, fig.width=10}
# Parameters for comprehensive visualization
path_plot_params <- list(
  v = 0.08,    # Moderate positive drift
  a = 1.0,
  z = 0.5,     # Unbiased
  s = 0.35,    # Moderate noise
  ter = 0.0,   # Use 0 for cleaner decision time plots
  dt = 0.01
)

n_example_paths <- 300 # Sufficient trials for good distributions

# Simulate the trials
set.seed(505)
example_trials_list <- vector("list", n_example_paths)
for (i in 1:n_example_paths) {
  example_trials_list[[i]] <- simulate_diffusion_trial_with_path(
    v = path_plot_params$v,
    a = path_plot_params$a,
    z = path_plot_params$z,
    s = path_plot_params$s,
    dt = path_plot_params$dt,
    ter = path_plot_params$ter
  )
}

# Check if we have enough valid trials
valid_trials <- sum(sapply(example_trials_list, function(x) !is.na(x$choice)))
cat("Generated", valid_trials, "valid trials out of", n_example_paths, "\n")

if(valid_trials > 50) {
  plot_ddm_paths_with_histograms(
    trials_data_list = example_trials_list,
    v_drift = path_plot_params$v,
    z_start = path_plot_params$z,
    a_threshold = path_plot_params$a,
    hist_binwidth = 0.15,
    max_time_to_plot = 6,
    main_plot_title = paste("DDM Simulation: Paths and RT Distributions\n",
                           "v =", path_plot_params$v, ", a =", path_plot_params$a,
                           ", z =", path_plot_params$z, ", s =", path_plot_params$s)
  )
} else {
  cat("Not enough valid trials with diverse outcomes to generate combined plot.\n")
}
```

**Interpretation**: - **Central panel**: Shows evidence accumulation paths over time - **Upper panel**: Distribution of decision times for upper boundary choices - **Lower panel**: Distribution of decision times for lower boundary choices (flipped) - **Orange line**: Expected drift trajectory based on parameter $v$ - **Dashed lines**: Decision boundaries and starting point

## Practical Applications and Model Fitting

### Simulating Different Experimental Conditions

The DDM can model various experimental manipulations:

```{r experimental_conditions, fig.height=6}
set.seed(2001)

# Define different experimental conditions
conditions <- list(
  "Easy High Accuracy" = list(v = 0.4, a = 1.0, z = 0.5),    # High drift
  "Hard Low Accuracy" = list(v = 0.1, a = 1.0, z = 0.5),     # Low drift  
  "Speed Emphasis" = list(v = 0.2, a = 0.8, z = 0.4),        # Low threshold
  "Accuracy Emphasis" = list(v = 0.2, a = 1.4, z = 0.7),     # High threshold
  "Response Bias" = list(v = 0.2, a = 1.0, z = 0.3)          # Biased start
)

# Simulate each condition
condition_results <- data.frame()
condition_data_list <- list()  # For plotting comparisons

for(cond_name in names(conditions)) {
  cond_params <- conditions[[cond_name]]
  sim_data <- simulate_diffusion_experiment(
    n_trials = 400,
    v = cond_params$v, 
    a = cond_params$a, 
    z = cond_params$z,
    s = 0.3, ter = 0.2, dt = 0.001
  )
  sim_data$condition <- cond_name
  condition_results <- rbind(condition_results, sim_data)
  condition_data_list[[cond_name]] <- sim_data
}

# Remove NA trials and calculate summaries
condition_summary <- condition_results %>%
  filter(!is.na(choice)) %>%
  group_by(condition) %>%
  summarise(
    n_trials = n(),
    accuracy = mean(choice == 1),  # Assuming upper boundary is "correct"
    mean_rt = mean(rt),
    median_rt = median(rt),
    rt_cv = sd(rt) / mean(rt),  # Coefficient of variation
    .groups = 'drop'
  )

knitr::kable(condition_summary, digits = 3, 
             caption = "Simulated Experimental Conditions")

# Visualize the speed-accuracy tradeoff
ggplot(condition_summary, aes(x = mean_rt, y = accuracy, label = condition)) +
  geom_point(size = 4, color = "steelblue") +
  geom_text(vjust = -0.8, hjust = 0.5, size = 3) +
  labs(title = "Speed-Accuracy Tradeoff Across Conditions",
       x = "Mean RT (s)", y = "Accuracy",
       subtitle = "Each point represents a different experimental condition") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"))
```

### Comparing RT Distributions Across Conditions

Using our utility function for comparing distributions:

```{r rt_comparison_plot, fig.height=6}
# Use the plotting utility to compare RT distributions
rt_comparison_plot <- plot_rt_comparison(
  condition_data_list[1:3],  # Compare first 3 conditions
  plot_type = "density",
  facet_by = "condition"
)

print(rt_comparison_plot)
```

### Model Validation: Predicted vs. Observed Patterns

```{r model_validation, fig.height=4}
# Focus on one condition for detailed analysis
focus_condition <- condition_results %>% 
  filter(condition == "Easy High Accuracy", !is.na(choice))

# Create RT quantile plots (a common DDM validation approach)
rt_quantiles <- focus_condition %>%
  group_by(choice) %>%
  summarise(
    q10 = quantile(rt, 0.1),
    q30 = quantile(rt, 0.3), 
    q50 = quantile(rt, 0.5),
    q70 = quantile(rt, 0.7),
    q90 = quantile(rt, 0.9),
    .groups = 'drop'
  ) %>%
  pivot_longer(cols = q10:q90, names_to = "quantile", values_to = "rt") %>%
  mutate(quantile_num = as.numeric(substr(quantile, 2, 3)) / 100)

ggplot(rt_quantiles, aes(x = quantile_num, y = rt, color = factor(choice))) +
  geom_point(size = 3) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = c("salmon", "steelblue"),
                     labels = c("Lower", "Upper"), name = "Boundary") +
  labs(title = "RT Quantiles by Choice (Model Validation)",
       x = "Quantile", y = "RT (s)",
       subtitle = "Shows the characteristic DDM pattern: faster responses for correct choices") +
  theme_minimal()
```

## Extensions and Advanced Topics

The basic DDM can be extended in several ways:

### 1. **Across-Trial Parameter Variability**

-   $sv$: Trial-to-trial drift rate variability
-   $sz$: Starting point variability\
-   $st$: Non-decision time variability

### 2. **Time-Varying Parameters**

-   Collapsing boundaries: $a(t) = a_0 \cdot e^{-\lambda t}$
-   Time-dependent drift: $v(t) = v_0 + v_1 \cdot t$

### 3. **Multiple Choice Extensions**

-   Racing diffusion models
-   Multi-alternative DDMs
-   Circular diffusion models

### 4. **Neural Implementations**

-   Linking DDM parameters to neural activity
-   Hierarchical drift-diffusion models
-   Neural DDMs with biophysical constraints

### 5. **Available Utility Functions**

Our enhanced DDM package now provides several utility functions for professional analysis:

**Parameter Management:** - `create_ddm_params()`: Create validated parameter sets - `create_parameter_grid()`: Generate systematic parameter combinations - `validate_ddm_parameters()`: Comprehensive parameter validation

**Data Analysis:** - `summarize_ddm_data()`: Comprehensive summary statistics - `create_summary_table()`: Formatted tables for reports

**Visualization:** - `plot_ddm_paths_with_histograms()`: Integrated path and RT visualization - `plot_rt_comparison()`: Compare RT distributions across conditions - `plot_rt_qq()`: QQ plots for distribution assessment

These functions make the DDM more accessible for research applications and provide professional-quality outputs for publications and presentations.

## Conclusion

This vignette demonstrated comprehensive simulation and analysis of the Diffusion Decision Model. Key takeaways:

1.  **Parameter Effects**: Each DDM parameter has distinct, interpretable effects on behavior
2.  **Speed-Accuracy Tradeoff**: The model naturally captures this fundamental decision-making phenomenon\
3.  **RT Distributions**: The DDM predicts characteristic right-skewed RT distributions
4.  **Flexibility**: The model can account for various experimental manipulations and individual differences
5.  **Professional Tools**: Enhanced utility functions provide systematic analysis capabilities

### What We've Covered

-   **Mathematical foundation**: Stochastic differential equation underlying the DDM
-   **Parameter exploration**: Systematic understanding of how each parameter affects behavior
-   **Simulation techniques**: From single trials to full experiments with path tracking
-   **Data analysis**: Comprehensive statistical summaries and visualization approaches
-   **Model validation**: QQ plots, quantile analysis, and distribution assessment
-   **Utility functions**: Professional tools for efficient DDM research

### Next Steps

In subsequent vignettes, we will explore: - **Parameter estimation**: Fitting DDM to empirical data - **Model comparison**: DDM vs. alternative decision models - **Advanced extensions**: Hierarchical and neural DDMs - **Real-world applications**: Clinical and cognitive assessment

### Recommended Readings

-   Ratcliff, R., & McKoon, G. (2008). The diffusion decision model: Theory and data. *Neural Computation*, 20(4), 873-922.
-   Wiecki, T. V., Sofer, I., & Frank, M. J. (2013). HDDM: Hierarchical Bayesian estimation of the drift-diffusion model in Python. *Frontiers in Neuroinformatics*, 7, 14.
-   Smith, P. L., & Ratcliff, R. (2004). Psychology and neurobiology of simple decisions. *Trends in Neurosciences*, 27(3), 161-168.

------------------------------------------------------------------------

*This tutorial provides a foundation for understanding and applying the DDM in cognitive research. The simulation approach allows researchers to build intuition about the model before applying it to empirical data. The enhanced utility functions ensure that your DDM analyses are both rigorous and professional.*
