---
title: "Understanding Parameter Variability in the DDM: From Fixed to Realistic Models"
description: "A comprehensive tutorial on across-trial parameter variability in the DDM"
order: 3
author: "Dogukan Nami Oztas"
date: "2025-05-16"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Understanding Parameter Variability in the DDM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup_variability, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = 'center',
  fig.width = 10,
  fig.height = 7
)

# Load necessary packages
library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)
library(scales)

# Source the DDM simulation functions
source("../R/02_ddm_simulator_basic.R")    # For standard DDM simulations
source("../R/03_ddm_simulator_variable.R") # For DDM simulations with variability
source("../R/utils/plot_parameter_comparison.R")

# Set consistent theme
theme_set(theme_minimal(base_size = 12))
```

## Introduction: Why Parameter Variability Matters

In the basic DDM introduced in the previous vignette, we assumed that core parameters like drift rate (*v*), starting point (*z*), and non-decision time (*ter*) remain constant across all trials within an experimental condition. While this simplification is useful for understanding the model's basic mechanics, it's **unrealistic** for modeling actual human behavior.

### The Problem with Fixed Parameters

Consider what happens in a real experiment:

- **Attention fluctuates**: Your focus isn't identical on trial 1 versus trial 500
- **Stimulus quality varies**: Even "identical" stimuli have subtle differences  
- **Processing efficiency changes**: You might get tired, distracted, or more practiced
- **Strategic adjustments occur**: You might become more or less cautious over time

These natural fluctuations mean that the "true" parameter values vary from trial to trial, even within the same experimental condition.

### The Solution: Across-Trial Parameter Variability

The **extended DDM** incorporates this reality by allowing key parameters to vary across trials according to specified distributions. This produces several important benefits:

1. **More realistic RT distributions**: Captures the full shape of empirical RT data
2. **Better error patterns**: Explains when errors are fast vs. slow  
3. **Improved model fit**: Accounts for variability that fixed-parameter models miss
4. **Richer theoretical insights**: Separates different sources of behavioral variation

## The Mathematics of Parameter Variability

### Core Idea: Parameters as Random Variables

Instead of fixed values, parameters become **random variables** drawn from distributions:

- **Drift rate**: $v_{trial} \sim \mathcal{N}(\mu_v, \sigma_v^2)$
- **Starting point**: $z_{trial} \sim \mathcal{U}(\mu_z - \frac{s_z}{2}, \mu_z + \frac{s_z}{2})$  
- **Non-decision time**: $ter_{trial} \sim \mathcal{U}(\mu_{ter} - \frac{s_{ter}}{2}, \mu_{ter} + \frac{s_{ter}}{2})$

Where:
- $\mu$ parameters are the **means** (what we estimated in the basic DDM)
- $\sigma_v$ (**sv**) is the **standard deviation** of drift rate variability
- $s_z$ (**sz**) is the **range** of starting point variability  
- $s_{ter}$ (**st0**) is the **range** of non-decision time variability

### Implementation in Our Simulator

Our `simulate_diffusion_trial_variable()` function implements this by:

1. **Sampling** trial-specific parameters from their distributions
2. **Running** the evidence accumulation with those values
3. **Recording** both the outcome and the parameters used

Let's see this in action!

## Part 1: Comparing Fixed vs. Variable Parameter Models

### Setting Up the Comparison

We'll start by comparing a basic DDM (fixed parameters) with a variable DDM using the same mean parameter values.

```{r setup_comparison}
# Shared simulation parameters
n_trials <- 3000  # Large enough to see distributional differences
base_params <- list(
  mean_v = 0.15,    # Moderate positive drift
  a = 0.7,          # Reasonable threshold
  mean_z = 0.35,     # Unbiased start (a/2)
  s = 0.3,         # Standard within-trial noise
  mean_ter = 0.08,  # Typical non-decision time
  dt = 0.001        # Fine time resolution
)

# Fixed DDM parameters (variability = 0)
fixed_params <- c(base_params, list(sv = 0, sz = 0, st0 = 0))

# Variable DDM parameters (realistic variability)
variable_params <- c(base_params, list(
  sv = 0.2,   # Moderate drift rate variability
  sz = 0.1,   # Some starting point variability  
  st0 = 0.04   # Small non-decision time variability
))

cat("Simulating Fixed DDM (sv = sz = st0 = 0)...")
set.seed(123)
data_fixed <- do.call(simulate_diffusion_experiment_variable, 
                     c(list(n_trials = n_trials), fixed_params))
data_fixed$model_type <- "Fixed Parameters"

cat("\nSimulating Variable DDM (sv = 0.12, sz = 0.08, st0 = 0.04)...")
set.seed(123) # Same seed for fair comparison
data_variable <- do.call(simulate_diffusion_experiment_variable,
                        c(list(n_trials = n_trials), variable_params))
data_variable$model_type <- "Variable Parameters"

cat("\nSimulations complete!")
```

### Visualizing the Dramatic Difference

```{r compare_fixed_vs_variable, fig.height=8}
# Combine data for plotting
combined_data <- bind_rows(
  data_fixed %>% select(rt, choice, model_type),
  data_variable %>% select(rt, choice, model_type)
) %>%
  filter(!is.na(rt), !is.na(choice)) %>%
  mutate(
    choice_label = factor(choice, levels = c(0, 1), 
                         labels = c("Lower Boundary", "Upper Boundary")),
    model_type = factor(model_type, levels = c("Fixed Parameters", "Variable Parameters"))
  )

# Create comprehensive comparison plot
p1 <- ggplot(combined_data, aes(x = rt, fill = model_type)) +
  geom_density(alpha = 0.7, color = "black", linewidth = 0.5) +
  facet_grid(model_type ~ choice_label, scales = "free_y") +
  scale_fill_manual(values = c("Fixed Parameters" = "#E74C3C", 
                              "Variable Parameters" = "#3498DB")) +
  #scale_x_continuous(limits = c(0, 1.5), breaks = seq(0, 1.5, 0.3)) +
  labs(
    title = "The Dramatic Impact of Parameter Variability",
    subtitle = "Same mean parameters, but very different RT distributions",
    x = "Reaction Time (seconds)",
    y = "Density",
    fill = "Model Type"
  ) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 11),
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12)
  )

print(p1)

# Calculate and display summary statistics
summary_stats <- combined_data %>%
  group_by(model_type, choice_label) %>%
  summarise(
    N = n(),
    Mean_RT = round(mean(rt), 3),
    Median_RT = round(median(rt), 3),
    SD_RT = round(sd(rt), 3),
    Skewness = round((mean(rt) - median(rt)) / sd(rt), 3),
    .groups = "drop"
  )

knitr::kable(summary_stats, 
             caption = "Key Statistics: Fixed vs. Variable Parameter Models")
```

### Key Observations

**Shape Transformation**: Parameter variability changes the RT distribution shape:

- **Fixed model**: Symmetric, narrow distributions
- **Variable model**: Right-skewed, wider distributions (more realistic!)

**Skewness**: The variable model produces the **positive skew** universally observed in real RT data, while the fixed model does not.

**Variability**: Real decisions show much more RT variability than fixed-parameter models predict.

## Part 2: Understanding Each Variability Component

Now let's systematically examine how each type of parameter variability affects the model's predictions.

### 2.1 Drift Rate Variability (sv): The Skewness Creator

Drift rate variability is often the **most important** variability parameter because it:
- Creates realistic right-skewed RT distributions
- Explains the relationship between correct and error RTs
- Accounts for trial-to-trial fluctuations in attention and processing

```{r drift_variability_analysis, fig.height=10}
# Parameters for drift rate variability exploration
sv_values <- c(0, 0.08, 0.16, 0.24)
n_trials_sv <- 2000

# Simulate across different sv values
sv_data_list <- list()
for (i in seq_along(sv_values)) {
  sv_val <- sv_values[i]
  set.seed(200 + i)
  
  params_sv <- c(base_params, list(sv = sv_val, sz = 0, st0 = 0))
  data_sv <- do.call(simulate_diffusion_experiment_variable,
                    c(list(n_trials = n_trials_sv), params_sv))
  
  data_sv$sv_condition <- paste0("sv = ", sv_val)
  data_sv$sv_value <- sv_val
  sv_data_list[[i]] <- data_sv
}

sv_combined <- bind_rows(sv_data_list) %>%
  filter(!is.na(rt), !is.na(choice)) %>%
  mutate(
    choice_label = factor(choice, levels = c(0, 1), 
                         labels = c("Lower Boundary", "Upper Boundary")),
    sv_condition = factor(sv_condition, levels = paste0("sv = ", sv_values))
  )

# Create the visualization
p_sv_density <- ggplot(sv_combined, aes(x = rt, fill = choice_label)) +
  geom_density(alpha = 0.7, color = "black", linewidth = 0.3) +
  facet_grid(sv_condition ~ choice_label) +
  scale_fill_manual(values = c("Lower Boundary" = "#E67E22", 
                              "Upper Boundary" = "#2ECC71")) +
  #scale_x_continuous(limits = c(0, 2), breaks = seq(0, 2, 0.5)) +
  labs(
    title = "Effect of Drift Rate Variability (sv) on RT Distributions",
    subtitle = "Higher sv creates more right-skewed distributions",
    x = "Reaction Time (seconds)",
    y = "Density",
    fill = "Response Type"
  ) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold"),
    plot.title = element_text(size = 14, face = "bold")
  )

# Calculate skewness measures
sv_skewness <- sv_combined %>%
  group_by(sv_condition, choice_label) %>%
  summarise(
    Mean_RT = round(mean(rt), 3),
    Median_RT = round(median(rt), 3),
    Skewness = round((mean(rt) - median(rt)) / sd(rt), 3),
    .groups = "drop"
  ) %>%
  select(sv_condition, choice_label, Skewness) %>%
  pivot_wider(names_from = choice_label, values_from = Skewness)

print(p_sv_density)
knitr::kable(sv_skewness, caption = "Skewness by sv Value")
```

**Key Insights about Drift Rate Variability:**

1. **Skewness increases with sv**: Higher values create more pronounced right tails
2. **Both responses affected**: sv influences the shape of both correct and error RT distributions  
3. **Realistic distributions**: sv ≥ 0.1 produces distributions similar to real data

### 2.2 Starting Point Variability (sz): The Fast Error Generator

Starting point variability primarily affects:
- The probability of fast errors (when you start close to the wrong boundary)
- Overall RT variance (but less dramatically than sv)
- Response bias patterns

```{r starting_point_variability, fig.height=8}
sz_values <- c(0, 0.05, 0.10, 0.15)
n_trials_sz <- 2000

# Simulate across different sz values  
sz_data_list <- list()
for (i in seq_along(sz_values)) {
  sz_val <- sz_values[i]
  set.seed(300 + i)
  
  params_sz <- c(base_params, list(sv = 0.1, sz = sz_val, st0 = 0))
  data_sz <- do.call(simulate_diffusion_experiment_variable,
                    c(list(n_trials = n_trials_sz), params_sz))
  
  data_sz$sz_condition <- paste0("sz = ", sz_val)
  data_sz$sz_value <- sz_val
  sz_data_list[[i]] <- data_sz
}

sz_combined <- bind_rows(sz_data_list) %>%
  filter(!is.na(rt), !is.na(choice)) %>%
  mutate(
    choice_label = factor(choice, levels = c(0, 1), 
                         labels = c("Lower Boundary", "Upper Boundary")),
    sz_condition = factor(sz_condition, levels = paste0("sz = ", sz_values))
  )

# Focus on fast errors - look at fastest 20% of responses
fast_errors_analysis <- sz_combined %>%
  group_by(sz_condition) %>%
  mutate(rt_percentile = ntile(rt, 5)) %>%
  filter(rt_percentile == 1) %>%  # Fastest 20%
  group_by(sz_condition, choice_label) %>%
  summarise(
    N_fast = n(),
    Prop_fast = n() / nrow(filter(sz_combined, sz_condition == first(sz_condition))),
    .groups = "drop"
  )

# Visualization focusing on leading edge
p_sz <- ggplot(sz_combined, aes(x = rt, fill = choice_label)) +
  geom_density(alpha = 0.7, color = "black", linewidth = 0.3) +
  facet_grid(sz_condition ~ choice_label) +
  scale_fill_manual(values = c("Lower Boundary" = "#E67E22", 
                              "Upper Boundary" = "#2ECC71")) +
  #scale_x_continuous(limits = c(0, 1.5), breaks = seq(0, 1.5, 0.3)) +
  labs(
    title = "Effect of Starting Point Variability (sz) on RT Distributions",
    subtitle = "Higher sz increases fast errors and overall variance",
    x = "Reaction Time (seconds)",
    y = "Density",
    fill = "Response Type"
  ) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold"),
    plot.title = element_text(size = 14, face = "bold")
  )

print(p_sz)
knitr::kable(fast_errors_analysis, 
             caption = "Fast Response Analysis: Effect of sz")
```

### 2.3 Non-Decision Time Variability (st0): The Leading Edge Smoother

Non-decision time variability has a distinctive effect:
- "Smears out" the leading edge of RT distributions
- Makes the fastest responses more variable
- Has relatively subtle effects compared to sv and sz

```{r nondecision_time_variability, fig.height=8}
st0_values <- c(0, 0.02, 0.04, 0.06)
n_trials_st0 <- 2000

# Simulate across different st0 values
st0_data_list <- list()
for (i in seq_along(st0_values)) {
  st0_val <- st0_values[i]
  set.seed(400 + i)
  
  params_st0 <- c(base_params, list(sv = 0.1, sz = 0, st0 = st0_val))
  data_st0 <- do.call(simulate_diffusion_experiment_variable,
                     c(list(n_trials = n_trials_st0), params_st0))
  
  data_st0$st0_condition <- paste0("st0 = ", st0_val)
  data_st0$st0_value <- st0_val
  st0_data_list[[i]] <- data_st0
}

st0_combined <- bind_rows(st0_data_list) %>%
  filter(!is.na(rt), !is.na(choice)) %>%
  mutate(
    choice_label = factor(choice, levels = c(0, 1), 
                         labels = c("Lower Boundary", "Upper Boundary")),
    st0_condition = factor(st0_condition, levels = paste0("st0 = ", st0_values))
  )

# Visualization with focus on leading edge
p_st0 <- ggplot(st0_combined, aes(x = rt, fill = choice_label)) +
  geom_density(alpha = 0.7, color = "black", linewidth = 0.3) +
  facet_grid(st0_condition ~ choice_label) +
  scale_fill_manual(values = c("Lower Boundary" = "#E67E22", 
                              "Upper Boundary" = "#2ECC71")) +
  #scale_x_continuous(limits = c(0, 1.2), breaks = seq(0, 1.2, 0.2)) +
  labs(
    title = "Effect of Non-Decision Time Variability (st0) on RT Distributions",
    subtitle = "Higher st0 smooths the leading edge of distributions",
    x = "Reaction Time (seconds)",
    y = "Density",
    fill = "Response Type"
  ) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold"),
    plot.title = element_text(size = 14, face = "bold")
  )

# Calculate leading edge statistics (10th percentile)
leading_edge_stats <- st0_combined %>%
  group_by(st0_condition, choice_label) %>%
  summarise(
    Q10 = round(quantile(rt, 0.1), 3),
    Q25 = round(quantile(rt, 0.25), 3),
    SD_RT = round(sd(rt), 3),
    .groups = "drop"
  )

print(p_st0)
knitr::kable(leading_edge_stats, 
             caption = "Leading Edge Analysis: Effect of st0")
```

## Part 3: The Combined Effects - Creating Realistic DDM Behavior

### Putting It All Together

Real behavior reflects the combined influence of all variability sources. Let's simulate a "realistic" DDM with all three types of variability:

```{r realistic_ddm_simulation, fig.height=10}
# Realistic parameter set
realistic_params <- list(
  mean_v = 0.2,   # Moderate drift rate
  a = 0.8,         # Reasonable threshold  
  mean_z = 0.4,   # Slightly biased start
  s = 0.2,         # Standard noise
  mean_ter = 0.18, # Typical non-decision time
  sv = 0.3,       # Substantial drift variability
  sz = 0.1,       # Moderate starting point variability
  st0 = 0.06,      # Small non-decision time variability
  dt = 0.001
)

# Simulate realistic DDM
set.seed(500)
realistic_data <- do.call(simulate_diffusion_experiment_variable,
                         c(list(n_trials = 3000), realistic_params))

# Also simulate a "textbook" version (no variability) for comparison
textbook_params <- realistic_params
textbook_params[c("sv", "sz", "st0")] <- 0

set.seed(500)
textbook_data <- do.call(simulate_diffusion_experiment_variable,
                        c(list(n_trials = 3000), textbook_params))

# Combine and prepare data
comparison_data <- bind_rows(
  realistic_data %>% mutate(model = "Realistic DDM\n(with variability)"),
  textbook_data %>% mutate(model = "Textbook DDM\n(no variability)")
) %>%
  filter(!is.na(rt), !is.na(choice)) %>%
  mutate(
    choice_label = factor(choice, levels = c(0, 1), 
                         labels = c("Error Responses", "Correct Responses")),
    model = factor(model, levels = c("Textbook DDM\n(no variability)", 
                                   "Realistic DDM\n(with variability)"))
  )

# Create comprehensive comparison
p_realistic <- ggplot(comparison_data, aes(x = rt, fill = model)) +
  geom_density(alpha = 0.7, color = "black", linewidth = 0.5) +
  facet_grid(model ~ choice_label) +
  scale_fill_manual(values = c("Textbook DDM\n(no variability)" = "#95A5A6",
                              "Realistic DDM\n(with variability)" = "#8E44AD")) +
  #scale_x_continuous(limits = c(0, 2), breaks = seq(0, 2, 0.4)) +
  labs(
    title = "Realistic vs. Textbook DDM: The Full Picture",
    subtitle = "Parameter variability creates distributions that match real human data",
    x = "Reaction Time (seconds)",
    y = "Density",
    fill = "Model Type"
  ) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 11),
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12)
  )

print(p_realistic)

# Statistical comparison
model_comparison <- comparison_data %>%
  group_by(model, choice_label) %>%
  summarise(
    N = n(),
    Mean_RT = round(mean(rt), 3),
    Median_RT = round(median(rt), 3),
    SD_RT = round(sd(rt), 3),
    Skewness = round((mean(rt) - median(rt)) / sd(rt), 3),
    Min_RT = round(min(rt), 3),
    Q90_RT = round(quantile(rt, 0.9), 3),
    .groups = "drop"
  )

knitr::kable(model_comparison, 
             caption = "Comprehensive Model Comparison")
```

### What Makes the Realistic DDM "Realistic"?

The realistic DDM captures several key features of real human RT data:

1. **Right-skewed distributions**: Long right tails for both correct and error responses
2. **Appropriate variability**: RT standard deviations that match empirical data
3. **Realistic leading edge**: Smoothed fastest responses, not sharp cutoffs
4. **Error patterns**: Mix of fast and slow errors, depending on trial-specific parameters

## Part 4: Understanding Trial-by-Trial Parameter Variation

### Examining the Sampled Parameters

Let's look at the actual parameter values sampled on each trial to understand the variability process:

```{r trial_by_trial_analysis, fig.height=8}
# Use the realistic simulation data which includes trial-specific parameters
param_analysis_data <- realistic_data %>%
  filter(!is.na(rt), !is.na(choice)) %>%
  mutate(
    choice_label = factor(choice, levels = c(0, 1), 
                         labels = c("Error", "Correct")),
    rt_fast = rt < quantile(rt, 0.25, na.rm = TRUE),
    rt_slow = rt > quantile(rt, 0.75, na.rm = TRUE)
  )

# Create panel plots showing parameter distributions
p1_params <- param_analysis_data %>%
  select(v_trial, z_trial, ter_trial, choice_label) %>%
  pivot_longer(cols = c(v_trial, z_trial, ter_trial), 
               names_to = "parameter", values_to = "value") %>%
  mutate(
    parameter = factor(parameter, 
                      levels = c("v_trial", "z_trial", "ter_trial"),
                      labels = c("Drift Rate (v)", "Starting Point (z)", "Non-Decision Time (ter)"))
  ) %>%
  ggplot(aes(x = value, fill = choice_label)) +
  geom_density(alpha = 0.7, color = "black", linewidth = 0.3) +
  facet_grid(parameter ~ choice_label, scales = "free_y") +
  scale_fill_manual(values = c("Error" = "#E74C3C", "Correct" = "#27AE60")) +
  labs(
    title = "Distribution of Trial-Specific Parameters",
    subtitle = "How parameter values vary across trials",
    x = "Parameter Value",
    y = "Density",
    fill = "Response Type"
  ) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold")
  )

print(p1_params)

# Examine parameter-RT relationships
rt_param_correlations <- param_analysis_data %>%
  summarise(
    cor_v_rt = round(cor(v_trial, rt, use = "complete.obs"), 3),
    cor_z_rt = round(cor(z_trial, rt, use = "complete.obs"), 3),
    cor_ter_rt = round(cor(ter_trial, rt, use = "complete.obs"), 3)
  )

cat("Correlations between trial parameters and RT:\n")
cat("v_trial - RT:", rt_param_correlations$cor_v_rt, "\n")
cat("z_trial - RT:", rt_param_correlations$cor_z_rt, "\n") 
cat("ter_trial - RT:", rt_param_correlations$cor_ter_rt, "\n")
```

### The Relationship Between Parameters and Outcomes

```{r parameter_outcome_relationships, fig.height=6}
# Create scatter plots showing parameter-RT relationships
p_relationships <- param_analysis_data %>%
  sample_n(1000) %>%  # Sample for clearer visualization
  select(rt, v_trial, z_trial, ter_trial, choice_label) %>%
  pivot_longer(cols = c(v_trial, z_trial, ter_trial), 
               names_to = "parameter", values_to = "value") %>%
  mutate(
    parameter = factor(parameter, 
                      levels = c("v_trial", "z_trial", "ter_trial"),
                      labels = c("Drift Rate (v)", "Starting Point (z)", "Non-Decision Time (ter)"))
  ) %>%
  ggplot(aes(x = value, y = rt, color = choice_label)) +
  geom_point(alpha = 0.6, size = 0.8) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1) +
  facet_wrap(~ parameter, scales = "free_x") +
  scale_color_manual(values = c("Error" = "#E74C3C", "Correct" = "#27AE60")) +
  scale_y_continuous(limits = c(0, 1.5)) +
  labs(
    title = "How Trial-Specific Parameters Influence RT",
    subtitle = "Each point represents one trial",
    x = "Parameter Value",
    y = "Reaction Time (seconds)",
    color = "Response Type"
  ) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold")
  )

print(p_relationships)
```

**Key Relationships:**

- **Drift rate (v)**: Higher values → faster RTs (stronger evidence accumulation)
- **Starting point (z)**: Variable effects depending on choice direction
- **Non-decision time (ter)**: Direct additive effect on total RT

## Part 5: Practical Implications and Guidelines

### When to Include Parameter Variability

**Always include** parameter variability when:
- You want realistic RT distributions that match human data
- You're fitting the DDM to empirical data
- You need to explain error RT patterns (fast vs. slow errors)
- You have sufficient data (>1000 trials per condition)

**Consider simplified models** when:
- Teaching basic DDM concepts
- You have very limited data (<200 trials)
- Computational efficiency is critical
- You're doing theoretical explorations

### Typical Parameter Values

Based on empirical DDM studies, typical variability parameter ranges are:

```{r parameter_guidelines, echo=FALSE}
guidelines <- data.frame(
  Parameter = c("sv (drift variability)", "sz (starting point variability)", 
                "st0 (non-decision time variability)"),
  `Typical Range` = c("0.05 - 0.30", "0.02 - 0.15", "0.01 - 0.10"),
  `Common Value` = c("0.10 - 0.15", "0.05 - 0.08", "0.03 - 0.05"),
  `Psychological Meaning` = c(
    "Attention fluctuations, stimulus variability",
    "Strategic adjustments, expectation changes", 
    "Motor preparation variability"
  )
)

knitr::kable(guidelines, caption = "Parameter Variability Guidelines")
```

### Model Selection Strategy

1. **Start simple**: Begin with fixed parameters to understand basic effects
2. **Add variability**: Include sv first (biggest impact), then sz and st0
3. **Compare models**: Use model comparison criteria (AIC, BIC) to justify complexity
4. **Validate**: Ensure parameter estimates are reasonable and stable

## Summary: The Power of Parameter Variability

Parameter variability transforms the DDM from a theoretical tool into a realistic model of human decision-making. The key insights are:

### Theoretical Insights
- **sv** (drift variability) is the primary driver of realistic RT distribution shapes
- **sz** (starting point variability) explains fast error patterns
- **st0** (non-decision time variability) accounts for motor and encoding fluctuations
- Combined effects create distributions that closely match human data

### Practical Implications  
- Parameter variability is essential for fitting real data
- Different variability sources have distinct signatures
- Model complexity should match data quantity and research goals
- Understanding variability helps interpret individual differences

### Conclusion

Across-trial parameter variability is a critical extension to the basic DDM. By allowing drift rate, starting point, and non-decision time to fluctuate from trial to trial, the model can more accurately capture the richness and complexity of human decision-making behavior, particularly the full shape of RT distributions and the patterns of error responses.

