---
title: "Understanding DDM Parameters: A Visual Guide to Decision-Making Components"
description: "An interactive exploration of how each DDM parameter shapes decisions and reaction times"
order: 4
author: "Dogukan Nami Oztas"
date: "2025-05-16"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 10
    fig_height: 7
vignette: >
  %\VignetteIndexEntry{Understanding DDM Parameters: A Visual Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = 'center',
  fig.width = 10,
  fig.height = 7,
  dev = 'png',
  dpi = 150
)

# Load necessary packages
library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)
library(viridis)
library(scales)
library(knitr)
library(stringr)
library(ggridges)

# Set consistent theme
theme_set(theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12, color = "gray40"),
    axis.title = element_text(size = 12),
    legend.position = "right",
    panel.grid.minor = element_blank()
  ))

# Source DDM functions
source("../R/03_ddm_simulator_variable.R")
source("../R/utils/plot_parameter_comparison.R")
source("../R/utils/plot_parameter_effects.R") # New enhanced plotting functions
```

## Learning Objectives

By the end of this tutorial, you will:

1.  **Understand** how each DDM parameter influences decision outcomes

2.  **Visualize** the effects of parameter changes on RT distributions and accuracy

3.  **Interpret** parameter values in cognitive terms

4.  **Predict** how parameter combinations affect behavior

5.  **Apply** this knowledge to design and interpret DDM studies

## Introduction: The Anatomy of a Decision

Imagine you're playing a simple video game where dots move across the screen, and you must quickly decide: are most dots moving left or right? This seemingly simple task involves multiple cognitive processes:

1.  **Evidence accumulation**: Your brain samples information about dot motion

2.  **Decision threshold**: You need "enough" evidence before committing

3.  **Starting bias**: You might expect one direction more than another

4.  **Motor execution**: Physical time to press the response button

The Diffusion Decision Model (DDM) elegantly captures these processes with just a few parameters. Let's explore each one!

```{r}
# Create a conceptual diagram of the DDM process
create_ddm_conceptual_diagram <- function() {
  # Create timeline data
  timeline_data <- data.frame(
    stage = c("Encoding", "Evidence\nAccumulation", "Motor\nResponse"),
    start = c(0, 0.1, 0.6),
    end = c(0.1, 0.6, 0.7),
    y = c(1, 1, 1),
    color = c("Non-decision", "Decision", "Non-decision")
  )
  
  # Create the plot
  ggplot(timeline_data) +
    geom_rect(aes(xmin = start, xmax = end, ymin = 0.8, ymax = 1.2, fill = color),
              alpha = 0.7) +
    geom_text(aes(x = (start + end)/2, y = y, label = stage),
              size = 5, fontface = "bold") +
    geom_segment(aes(x = 0, xend = 0.7, y = 0.5, yend = 0.5),
                 arrow = arrow(length = unit(0.3, "cm")),
                 size = 1.5) +
    scale_x_continuous(breaks = c(0, 0.1, 0.6, 0.7),
                       labels = c("0", "ter", "ter + DT", "RT"),
                       limits = c(-0.05, 0.75)) +
    scale_fill_manual(values = c("Decision" = "steelblue", 
                                 "Non-decision" = "gray70")) +
    labs(title = "Components of Response Time in the DDM",
         subtitle = "RT = Non-decision time (ter) + Decision time (DT)",
         x = "Time (seconds)",
         y = "") +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          panel.grid = element_blank(),
          legend.title = element_blank())
}

create_ddm_conceptual_diagram()
```

## Setting Up Our Baseline Model

Let's establish a baseline set of parameters that we'll use throughout this tutorial:

```{r}
# Define baseline parameters with clear documentation
baseline_params <- list(
  # Core parameters
  mean_v = 0.15,      # Moderate positive drift (slightly favors upper boundary)
  a = 0.7,            # Moderate threshold (balance of speed/accuracy)
  mean_z = 0.35,       # Unbiased starting point (middle of threshold)
  
  # Noise and timing
  s = 0.25,            # Within-trial noise (standard for scaling)
  mean_ter = 0.2,     # 200ms non-decision time
  
  # Variability parameters (set to 0 for baseline)
  sv = 0,             # No drift variability
  sz = 0,             # No starting point variability
  st0 = 0,            # No non-decision time variability
  
  # Simulation parameters
  dt = 0.001,         # High precision time step
  n_trials = 1000     # Sufficient for stable distributions
)

# Create a summary table
param_table <- data.frame(
  Parameter = c("Drift rate (v)", "Threshold (a)", "Starting point (z)", 
                "Noise (s)", "Non-decision time (ter)"),
  Symbol = c("v", "a", "z", "s", "ter"),
  Baseline = c(0.15, 1.0, 0.5, 0.1, 0.2),
  Interpretation = c(
    "Quality of evidence (positive = favors upper)",
    "Amount of evidence required",
    "Initial bias (0.5 = unbiased)",
    "Moment-to-moment variability",
    "Time for encoding + motor response"
  )
)

kable(param_table, caption = "Baseline DDM Parameters")
```

Now let's simulate our baseline model:

```{r}
set.seed(2024)  # For reproducibility

# Simulate baseline data
baseline_data <- do.call(
  simulate_diffusion_experiment_variable,
  c(baseline_params)
)

# Add descriptive labels
baseline_data$condition <- "Baseline"
baseline_data$param_set <- "Baseline"

# Summary statistics
baseline_summary <- baseline_data %>%
  filter(!is.na(choice)) %>%
  summarise(
    n_valid = n(),
    prop_upper = mean(choice == 1),
    mean_rt = mean(rt),
    median_rt = median(rt),
    rt_10 = quantile(rt, 0.1),
    rt_90 = quantile(rt, 0.9),
    .groups = 'drop'
  )

cat("Baseline Model Summary:\n")
cat("- Valid trials:", baseline_summary$n_valid, "\n")
cat("- Upper boundary choices:", percent(baseline_summary$prop_upper), "\n")
cat("- Mean RT:", round(baseline_summary$mean_rt, 3), "s\n")
cat("- RT range (10th-90th percentile):", 
    round(baseline_summary$rt_10, 3), "-", 
    round(baseline_summary$rt_90, 3), "s\n")
```

## Part 1: Drift Rate (v) - The Engine of Evidence

The drift rate represents how quickly and in which direction evidence accumulates. Think of it as the "signal strength" in your decision.

### Cognitive Interpretation

-   **Magnitude (\|v\|)**: Task difficulty or stimulus quality

    -   High \|v\| → Clear evidence → Fast, accurate decisions

    -   Low \|v\| → Ambiguous evidence → Slow, error-prone decisions

-   **Sign**: Direction of evidence

    -   Positive v → Evidence favors upper boundary

    -   Negative v → Evidence favors lower boundary

    -   v = 0 → No systematic evidence (pure guessing)

### Exploring Drift Rate Effects

```{r}
# Define drift rate values to explore
drift_values <- c(-0.3, -0.1, 0, 0.1, 0.3)
drift_labels <- c("Strong Lower", "Weak Lower", "No Drift", 
                  "Weak Upper", "Strong Upper")

# Simulate data for different drift rates
drift_data_list <- list()

for (i in seq_along(drift_values)) {
  params <- baseline_params
  params$mean_v <- drift_values[i]
  
  set.seed(2024 + i)
  sim_data <- do.call(
    simulate_diffusion_experiment_variable,
    c(params)
  )
  
  sim_data$drift_rate <- drift_values[i]
  sim_data$drift_label <- factor(drift_labels[i], levels = drift_labels)
  drift_data_list[[i]] <- sim_data
}

drift_data <- bind_rows(drift_data_list)

# Create comprehensive visualization
p1 <- ggplot(drift_data %>% filter(!is.na(choice)), 
             aes(x = drift_label, fill = factor(choice))) +
  geom_bar(position = "fill", alpha = 0.8) +
  scale_fill_manual(values = c("0" = "coral", "1" = "steelblue"),
                    labels = c("Lower", "Upper")) +
  scale_y_continuous(labels = percent) +
  labs(title = "A. Choice Proportions by Drift Rate",
       subtitle = "Higher drift → More upper boundary choices",
       x = "Drift Rate Condition",
       y = "Proportion of Choices",
       fill = "Boundary") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p2 <- ggplot(drift_data %>% filter(!is.na(choice)), 
             aes(x = rt, fill = drift_label)) +
  geom_density(alpha = 0.6) +
  facet_wrap(~ factor(choice, labels = c("Lower Boundary", "Upper Boundary"))) +
  scale_fill_viridis_d() +
  #coord_cartesian(xlim = c(0, 2)) +
  labs(title = "B. RT Distributions by Drift Rate",
       subtitle = "Stronger drift → Faster decisions",
       x = "Reaction Time (s)",
       y = "Density",
       fill = "Drift Condition")

# Summary statistics plot
drift_summary <- drift_data %>%
  filter(!is.na(choice)) %>%
  group_by(drift_rate, drift_label) %>%
  summarise(
    accuracy = mean(choice == (drift_rate > 0)),
    mean_rt = mean(rt),
    .groups = 'drop'
  )

p3 <- ggplot(drift_summary, aes(x = drift_rate)) +
  geom_line(aes(y = accuracy), color = "darkgreen", size = 2) +
  geom_point(aes(y = accuracy), color = "darkgreen", size = 4) +
  scale_y_continuous(labels = percent, limits = c(0, 1)) +
  labs(title = "C. Accuracy vs Drift Rate",
       subtitle = "Psychometric function",
       x = "Drift Rate (v)",
       y = "Accuracy")

p4 <- ggplot(drift_summary, aes(x = drift_rate)) +
  geom_line(aes(y = mean_rt), color = "darkblue", size = 2) +
  geom_point(aes(y = mean_rt), color = "darkblue", size = 4) +
  labs(title = "D. Mean RT vs Drift Rate",
       subtitle = "Chronometric function",
       x = "Drift Rate (v)",
       y = "Mean RT (s)")

# Combine plots
(p1 + p2) / (p3 + p4) +
  plot_annotation(
    title = "Effects of Drift Rate on Decision Behavior",
    theme = theme(plot.title = element_text(size = 18, face = "bold"))
  )
```

### Key Insights: Drift Rate

1.  **Choice Probability**: Follows a sigmoid (S-shaped) curve

2.  **Reaction Times**: Decrease with increasing \|v\|

3.  **Accuracy**: Approaches chance (50%) as v → 0

4.  **Symmetry**: Effects are symmetric around v = 0

## Part 2: Threshold Separation (a) - The Speed-Accuracy Tradeoff

The threshold separation controls how much evidence is required before making a decision. It's the primary parameter controlling the speed-accuracy tradeoff.

### Cognitive Interpretation

-   **High threshold (large a)**: Conservative strategy

    -   More evidence required → Slower but more accurate

    -   "I want to be sure before I respond"

-   **Low threshold (small a)**: Liberal strategy

    -   Less evidence required → Faster but more error-prone

    -   "I need to respond quickly"

```{r}
# Define threshold values to explore
threshold_values <- c(0.5, 0.75, 1.0, 1.25, 1.5)
threshold_labels <- c("Very Liberal", "Liberal", "Balanced", 
                      "Conservative", "Very Conservative")

# Simulate data for different thresholds
threshold_data_list <- list()

for (i in seq_along(threshold_values)) {
  params <- baseline_params
  params$a <- threshold_values[i]
  params$mean_z <- threshold_values[i] / 2  # Keep unbiased
  
  set.seed(3024 + i)
  sim_data <- do.call(
    simulate_diffusion_experiment_variable,
    c(params)
  )
  
  sim_data$threshold <- threshold_values[i]
  sim_data$threshold_label <- factor(threshold_labels[i], levels = threshold_labels)
  threshold_data_list[[i]] <- sim_data
}

threshold_data <- bind_rows(threshold_data_list)

# Create visualizations
# Speed-Accuracy Tradeoff plot
threshold_summary <- threshold_data %>%
  filter(!is.na(choice)) %>%
  group_by(threshold, threshold_label) %>%
  summarise(
    accuracy = mean(choice == 1),  # Since v > 0
    mean_rt = mean(rt),
    rt_variability = sd(rt),
    .groups = 'drop'
  )

p1 <- ggplot(threshold_summary, aes(x = mean_rt, y = accuracy)) +
  geom_path(color = "gray50", size = 1, alpha = 0.5) +
  geom_point(aes(color = threshold), size = 5) +
  geom_text(aes(label = threshold_label), 
            vjust = -1, hjust = 0.5, size = 3) +
  scale_color_viridis_c(name = "Threshold (a)") +
  #scale_y_continuous(labels = percent, limits = c(0.5, 1)) +
  labs(title = "A. The Speed-Accuracy Tradeoff",
       subtitle = "Higher thresholds → Slower but more accurate responses",
       x = "Mean RT (s)",
       y = "Accuracy")

# RT distributions
p2 <- ggplot(threshold_data %>% filter(!is.na(choice)), 
             aes(x = rt, fill = threshold_label)) +
  geom_density(alpha = 0.6) +
  scale_fill_viridis_d() +
  #coord_cartesian(xlim = c(0, 3)) +
  labs(title = "B. RT Distributions by Threshold",
       subtitle = "Higher thresholds → More spread out distributions",
       x = "Reaction Time (s)",
       y = "Density",
       fill = "Threshold")

# RT variability
p3 <- ggplot(threshold_summary, aes(x = threshold, y = rt_variability)) +
  geom_line(color = "darkred", size = 2) +
  geom_point(color = "darkred", size = 4) +
  labs(title = "C. RT Variability vs Threshold",
       subtitle = "Higher thresholds → More variable RTs",
       x = "Threshold (a)",
       y = "RT Standard Deviation (s)")

# Error RT analysis
error_analysis <- threshold_data %>%
  filter(!is.na(choice)) %>%
  group_by(threshold, threshold_label, choice) %>%
  summarise(
    mean_rt = mean(rt),
    .groups = 'drop'
  ) %>%
  mutate(response_type = ifelse(choice == 1, "Correct", "Error"))

p4 <- ggplot(error_analysis, aes(x = threshold, y = mean_rt, 
                                  color = response_type)) +
  geom_line(size = 2) +
  geom_point(size = 4) +
  scale_color_manual(values = c("Correct" = "darkgreen", "Error" = "darkred")) +
  labs(title = "D. Correct vs Error RTs",
       subtitle = "Pattern changes with threshold",
       x = "Threshold (a)",
       y = "Mean RT (s)",
       color = "Response Type")

# Combine plots
(p1 + p2) / (p3 + p4) +
  plot_annotation(
    title = "Effects of Threshold on Decision Behavior",
    theme = theme(plot.title = element_text(size = 18, face = "bold"))
  )
```

### Key Insights: Threshold

1.  **Speed-Accuracy Tradeoff**: Fundamental relationship in decision-making

2.  **RT Distributions**: Become wider and more skewed with higher thresholds

3.  **Error Patterns**: Can shift between fast and slow errors

4.  **Strategic Control**: Reflects cognitive control and task demands

## Part 3: Starting Point (z) - Decision Bias

The starting point determines where evidence accumulation begins, creating a bias toward one response.

### Cognitive Interpretation

-   **z = a/2**: Unbiased (equidistant from both boundaries)

-   **z \> a/2**: Biased toward upper boundary

-   **z \< a/2**: Biased toward lower boundary

This can reflect:

-   Prior expectations ("I expect to see motion to the right")

-   Reward asymmetries ("Correct responses pay more than errors")

-   Strategic biases ("When unsure, guess 'yes'")

```{r}
# Define starting point values
z_proportions <- c(0.2, 0.35, 0.5, 0.65, 0.8)
z_labels <- c("Strong Lower Bias", "Moderate Lower Bias", "Unbiased", 
              "Moderate Upper Bias", "Strong Upper Bias")

# Simulate data
z_data_list <- list()

for (i in seq_along(z_proportions)) {
  params <- baseline_params
  params$mean_z <- z_proportions[i] * params$a
  
  set.seed(4024 + i)
  sim_data <- do.call(
    simulate_diffusion_experiment_variable,
    c(params)
  )
  
  sim_data$z_prop <- z_proportions[i]
  sim_data$z_label <- factor(z_labels[i], levels = z_labels)
  z_data_list[[i]] <- sim_data
}

z_data <- bind_rows(z_data_list)

# Visualizations
# Choice proportions
z_summary <- z_data %>%
  filter(!is.na(choice)) %>%
  group_by(z_prop, z_label) %>%
  summarise(
    prop_upper = mean(choice == 1),
    mean_rt_upper = mean(rt[choice == 1]),
    mean_rt_lower = mean(rt[choice == 0]),
    .groups = 'drop'
  )

p1 <- ggplot(z_summary, aes(x = z_prop, y = prop_upper)) +
  geom_line(color = "purple", size = 2) +
  geom_point(color = "purple", size = 4) +
  geom_hline(yintercept = 0.5, linetype = "dashed", alpha = 0.5) +
  scale_y_continuous(labels = percent, limits = c(0, 1)) +
  scale_x_continuous(labels = percent) +
  labs(title = "A. Choice Probability vs Starting Point",
       subtitle = "Bias creates preference for closer boundary",
       x = "Starting Point (z/a)",
       y = "P(Upper Boundary)")

# RT by response
z_rt_data <- z_data %>%
  filter(!is.na(choice)) %>%
  group_by(z_prop, z_label, choice) %>%
  summarise(mean_rt = mean(rt), .groups = 'drop') %>%
  mutate(response = ifelse(choice == 1, "Upper", "Lower"))

p2 <- ggplot(z_rt_data, aes(x = z_prop, y = mean_rt, color = response)) +
  geom_line(size = 2) +
  geom_point(size = 4) +
  scale_color_manual(values = c("Upper" = "steelblue", "Lower" = "coral")) +
  scale_x_continuous(labels = percent) +
  labs(title = "B. Mean RT by Response and Bias",
       subtitle = "Responses toward biased boundary are faster",
       x = "Starting Point (z/a)",
       y = "Mean RT (s)",
       color = "Response")

# Detailed RT distributions for extreme biases
p3 <- z_data %>%
  filter(!is.na(choice), z_prop %in% c(0.2, 0.5, 0.8)) %>%
  ggplot(aes(x = rt, fill = factor(choice))) +
  geom_density(alpha = 0.6) +
  facet_grid(z_label ~ ., scales = "free_y") +
  scale_fill_manual(values = c("0" = "coral", "1" = "steelblue"),
                    labels = c("Lower", "Upper")) +
  #coord_cartesian(xlim = c(0, 2)) +
  labs(title = "C. RT Distributions Under Different Biases",
       subtitle = "Bias affects both speed and frequency of responses",
       x = "Reaction Time (s)",
       y = "Density",
       fill = "Boundary")

# Bias effect visualization
p4 <- expand.grid(
  z_prop = seq(0.1, 0.9, 0.01),
  boundary = c("Lower", "Upper")
) %>%
  mutate(
    distance = ifelse(boundary == "Upper", 1 - z_prop, z_prop),
    expected_rt = 0.2 + distance * 2  # Simplified model
  ) %>%
  ggplot(aes(x = z_prop, y = expected_rt, color = boundary)) +
  geom_line(size = 2) +
  scale_color_manual(values = c("Upper" = "steelblue", "Lower" = "coral")) +
  scale_x_continuous(labels = percent) +
  labs(title = "D. Theoretical RT vs Starting Point",
       subtitle = "Distance to boundary determines speed",
       x = "Starting Point (z/a)",
       y = "Expected RT (arbitrary units)",
       color = "Boundary")

# Combine
(p1 + p2) / (p3 + p4) +
  plot_annotation(
    title = "Effects of Starting Point Bias",
    theme = theme(plot.title = element_text(size = 18, face = "bold"))
  )
```

## Part 4: Variability Parameters - Real-World Complexity

Real decisions aren't perfectly consistent. The DDM includes parameters to capture trial-to-trial variability:

### Types of Variability

1.  **Drift variability (sv)**: Attention fluctuations, stimulus differences

2.  **Starting point variability (sz)**: Expectation uncertainty

3.  **Non-decision time variability (st0)**: Motor preparation differences

```{r}
# Compare models with and without drift variability
params_no_sv <- baseline_params
params_with_sv <- baseline_params
params_with_sv$sv <- 0.15

set.seed(5024)
data_no_sv <- do.call(
  simulate_diffusion_experiment_variable,
  c(params_no_sv)
)
data_no_sv$model <- "No Variability (sv = 0)"

data_with_sv <- do.call(
  simulate_diffusion_experiment_variable,
  c(params_with_sv)
)
data_with_sv$model <- "With Variability (sv = 0.15)"

variability_data <- bind_rows(data_no_sv, data_with_sv)

# Key effect: RT distributions become more skewed
p1 <- variability_data %>%
  filter(!is.na(choice)) %>%
  ggplot(aes(x = rt, fill = model)) +
  geom_density(alpha = 0.6) +
  facet_wrap(~ factor(choice, labels = c("Errors", "Correct"))) +
  scale_fill_manual(values = c("blue", "red")) +
  #coord_cartesian(xlim = c(0, 3)) +
  labs(title = "A. Effect of Drift Variability on RT Distributions",
       subtitle = "sv > 0 creates heavier tails, especially for errors",
       x = "Reaction Time (s)",
       y = "Density",
       fill = "Model")

# Quantile-quantile plot
qq_data <- variability_data %>%
  filter(!is.na(choice)) %>%
  group_by(model, choice) %>%
  summarise(
    q10 = quantile(rt, 0.1),
    q30 = quantile(rt, 0.3),
    q50 = quantile(rt, 0.5),
    q70 = quantile(rt, 0.7),
    q90 = quantile(rt, 0.9),
    .groups = 'drop'
  ) %>%
  pivot_longer(cols = starts_with("q"), 
               names_to = "quantile", 
               values_to = "rt") %>%
  mutate(quantile_num = as.numeric(str_extract(quantile, "\\d+")) / 100)

p2 <- ggplot(qq_data, aes(x = quantile_num, y = rt, 
                          color = model, linetype = factor(choice))) +
  geom_line(size = 1.5) +
  geom_point(size = 3) +
  scale_color_manual(values = c("blue", "red")) +
  scale_linetype_manual(values = c("dashed", "solid"),
                        labels = c("Errors", "Correct")) +
  labs(title = "B. Quantile Functions",
       subtitle = "sv affects the spread of quantiles differently for correct/error responses",
       x = "Quantile",
       y = "RT (s)",
       color = "Model",
       linetype = "Response")

# Show trial-by-trial variability
trial_params <- data_with_sv %>%
  slice(1:100) %>%
  select(trial, v_trial, z_trial, ter_trial)

p3 <- trial_params %>%
  pivot_longer(cols = -trial, names_to = "parameter", values_to = "value") %>%
  mutate(parameter = factor(parameter, 
                           levels = c("v_trial", "z_trial", "ter_trial"),
                           labels = c("Drift (v)", "Starting Point (z)", "Non-decision (ter)"))) %>%
  ggplot(aes(x = trial, y = value)) +
  geom_line(color = "darkgreen") +
  geom_point(size = 1, alpha = 0.5) +
  facet_wrap(~ parameter, scales = "free_y", ncol = 1) +
  labs(title = "C. Trial-by-Trial Parameter Values",
       subtitle = "Example of parameter variability across 100 trials",
       x = "Trial Number",
       y = "Parameter Value")

# Summary of variability effects
var_summary <- data.frame(
  Parameter = c("sv (Drift)", "sz (Starting Point)", "st0 (Non-decision)"),
  `Primary Effect` = c(
    "Increases RT distribution skew",
    "Can create fast errors",
    "Smooths leading edge of RT distribution"
  ),
  `Cognitive Interpretation` = c(
    "Attention fluctuations, stimulus variability",
    "Expectation uncertainty, preparation variability",
    "Motor readiness, encoding variability"
  )
)

p4 <- ggplot() +
  annotation_custom(
    tableGrob(var_summary, rows = NULL,
              theme = ttheme_minimal(base_size = 10)),
    xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf
  ) +
  theme_void() +
  labs(title = "D. Summary of Variability Parameters")

# Combine
(p1 + p2) / (p3 + p4) +
  plot_annotation(
    title = "The Role of Trial-to-Trial Variability",
    theme = theme(plot.title = element_text(size = 18, face = "bold"))
  )
```

## Part 5: Parameter Interactions

Parameters don't work in isolation. Let's explore some important interactions:

```{r}
# Create a grid of parameter combinations
interaction_grid <- expand.grid(
  drift = c(0.05, 0.25),
  threshold = c(0.7, 1, 1.3),
  bias = c(0.3, 0.5, 0.7)
) %>%
  mutate(
    condition = paste0("v=", drift, ", a=", threshold, ", z/a=", bias)
  )

# Simulate data for each combination
interaction_data_list <- list()

for (i in 1:nrow(interaction_grid)) {
  params <- baseline_params
  params$mean_v <- interaction_grid$drift[i]
  params$a <- interaction_grid$threshold[i]
  params$mean_z <- interaction_grid$bias[i] * interaction_grid$threshold[i]
  
  set.seed(6024 + i)
  sim_data <- do.call(
   simulate_diffusion_experiment_variable,
   c(params)
 )
 
 sim_data$drift <- interaction_grid$drift[i]
 sim_data$threshold <- interaction_grid$threshold[i]
 sim_data$bias <- interaction_grid$bias[i]
 sim_data$condition <- interaction_grid$condition[i]
 interaction_data_list[[i]] <- sim_data
}

interaction_data <- bind_rows(interaction_data_list)

# Analyze interactions
interaction_summary <- interaction_data %>%
 filter(!is.na(choice)) %>%
 group_by(drift, threshold, bias) %>%
 summarise(
   prop_upper = mean(choice == 1),
   mean_rt = mean(rt),
   accuracy = mean(choice == (bias >= 0.5)),  # Simplified accuracy measure
   .groups = 'drop'
 )

# Visualization 1: Drift × Threshold interaction
p1 <- interaction_summary %>%
 filter(bias == 0.5) %>%  # Unbiased only
 ggplot(aes(x = factor(drift), y = mean_rt, fill = factor(threshold))) +
 geom_bar(stat = "identity", position = "dodge", alpha = 0.8) +
 scale_fill_viridis_d(name = "Threshold") +
 labs(title = "A. Drift vs Threshold interaction on RT",
      subtitle = "Effect of drift is amplified with higher thresholds",
      x = "Drift Rate",
      y = "Mean RT (s)")

# Visualization 2: Bias × Drift interaction
p2 <- interaction_summary %>%
 filter(threshold == 1) %>%  # Middle threshold
 ggplot(aes(x = bias, y = prop_upper, color = factor(drift))) +
 geom_line(size = 2) +
 geom_point(size = 4) +
 scale_color_viridis_d(name = "Drift") +
 scale_x_continuous(labels = percent) +
 scale_y_continuous(labels = percent) +
 labs(title = "B. Bias vs Drift Interaction on Choice",
      subtitle = "Bias effect is stronger with weak drift",
      x = "Starting Point Bias (z/a)",
      y = "P(Upper Boundary)")

# Visualization 3: Three-way interaction heatmap
p3 <- interaction_summary %>%
 ggplot(aes(x = factor(drift), y = factor(bias), fill = mean_rt)) +
 geom_tile() +
 facet_wrap(~ paste("Threshold =", threshold)) +
 scale_fill_viridis_c(name = "Mean RT") +
 labs(title = "C. Three-Way Interaction: RT Patterns",
      subtitle = "Complex interplay between all parameters",
      x = "Drift Rate",
      y = "Starting Bias")

# Visualization 4: Interactive effects on accuracy
p4 <- interaction_summary %>%
 ggplot(aes(x = threshold, y = accuracy, color = factor(drift))) +
 geom_line(size = 2) +
 geom_point(size = 4) +
 facet_wrap(~ paste("Bias =", bias)) +
 scale_color_viridis_d(name = "Drift") +
 scale_y_continuous(labels = percent) +
 labs(title = "D. Parameter Effects on Accuracy",
      subtitle = "Accuracy depends on parameter combinations",
      x = "Threshold",
      y = "Accuracy")

# Combine
(p1 + p2) / (p3 + p4) +
 plot_annotation(
   title = "Parameter Interactions in the DDM",
   theme = theme(plot.title = element_text(size = 18, face = "bold"))
 )
```

## Interactive Exploration

Let's create an interactive tool to explore parameters in real-time:

```{r}
# Create a function for interactive parameter exploration
explore_ddm_parameter <- function(param_name, param_values, param_labels = NULL) {
  if (is.null(param_labels)) {
    param_labels <- as.character(param_values)
  }
  
  # Simulate data for each parameter value
  explore_data_list <- list()
  
  for (i in seq_along(param_values)) {
    params <- baseline_params
    
    # Set the parameter of interest
    if (param_name == "v") {
      params$mean_v <- param_values[i]
    } else if (param_name == "a") {
      params$a <- param_values[i]
      params$mean_z <- param_values[i] / 2  # Keep unbiased
    } else if (param_name == "z_rel") {
      params$mean_z <- param_values[i] * params$a
    } else if (param_name == "ter") {
      params$mean_ter <- param_values[i]
    }
    
    set.seed(7024 + i)
    sim_data <- do.call(
      simulate_diffusion_experiment_variable,
      c(params)
    )
    
    sim_data$param_value <- param_values[i]
    sim_data$param_label <- param_labels[i]
    explore_data_list[[i]] <- sim_data
  }
  
  explore_data <- bind_rows(explore_data_list)
  
  # Create visualization
  p1 <- explore_data %>%
    filter(!is.na(choice)) %>%
    ggplot(aes(x = rt, y = param_label, fill = factor(choice))) +
    geom_density_ridges(alpha = 0.7, scale = 0.9) +
    scale_fill_manual(values = c("0" = "coral", "1" = "steelblue"),
                      labels = c("Lower", "Upper")) +
    #coord_cartesian(xlim = c(0, 2)) +
    labs(title = paste("RT Distributions by", param_name),
         x = "Reaction Time (s)",
         y = "Parameter Value",
         fill = "Boundary")
  
  # Summary metrics
  summary_data <- explore_data %>%
    filter(!is.na(choice)) %>%
    group_by(param_value, param_label) %>%
    summarise(
      prop_upper = mean(choice == 1),
      mean_rt = mean(rt),
      rt_skew = (mean(rt) - median(rt)) / sd(rt),
      .groups = 'drop'
    )
  
  p2 <- summary_data %>%
    pivot_longer(cols = c(prop_upper, mean_rt, rt_skew),
                 names_to = "metric", values_to = "value") %>%
    mutate(metric = factor(metric, 
                          levels = c("prop_upper", "mean_rt", "rt_skew"),
                          labels = c("P(Upper)", "Mean RT", "RT Skew"))) %>%
    ggplot(aes(x = param_value, y = value, color = metric)) +
    geom_line(size = 2) +
    geom_point(size = 4) +
    facet_wrap(~ metric) +
    scale_color_viridis_d() +
    labs(title = paste("Summary Metrics by", param_name),
         x = "Parameter Value",
         y = "Metric Value") +
    theme(legend.position = "none")
  
  return(p1 / p2)
}

# Example: Explore non-decision time
ter_plot <- explore_ddm_parameter(
  param_name = "ter",
  param_values = c(0.1, 0.2, 0.3, 0.4),
  param_labels = c("100ms", "200ms", "300ms", "400ms")
) +
  plot_annotation(
    title = "Effect of Non-Decision Time",
    subtitle = "ter shifts the entire RT distribution without changing its shape",
    theme = theme(plot.title = element_text(size = 16, face = "bold"))
  )

print(ter_plot)
```

## Practical Guidelines

Based on our exploration, here are practical guidelines for interpreting DDM parameters:

```{r}
# Create a reference guide
guidelines <- data.frame(
  Parameter = c("Drift Rate (v)", "Threshold (a)", "Starting Point (z)", 
                "Drift Variability (sv)", "Non-decision Time (ter)"),
  `Typical Range` = c("[-0.5, 0.5]", "[0.5, 2.0]", "[0.3a, 0.7a]", 
                      "[0, 0.3]", "[0.1, 0.5]"),
  `What High Values Mean` = c(
    "Strong/clear evidence",
    "Cautious/accurate responding",
    "Bias toward upper boundary",
    "Variable attention/difficulty",
    "Slow encoding/motor processes"
  ),
  `What Low Values Mean` = c(
    "Weak/ambiguous evidence",
    "Fast/risky responding",
    "Bias toward lower boundary",
    "Consistent processing",
    "Fast encoding/motor processes"
  ),
  `Key Diagnostic` = c(
    "Accuracy & mean RT",
    "Speed-accuracy tradeoff",
    "Choice proportions",
    "RT distribution skew",
    "Minimum RT"
  )
)

kable(guidelines, caption = "Quick Reference: DDM Parameter Interpretation")

# Create a diagnostic plot function
create_diagnostic_plot <- function(data, title = "DDM Diagnostics") {
  # Calculate key metrics
  diagnostics <- data %>%
    filter(!is.na(choice)) %>%
    summarise(
      n_trials = n(),
      prop_upper = mean(choice == 1),
      mean_rt = mean(rt),
      min_rt = min(rt),
      rt_skew = (mean(rt) - median(rt)) / sd(rt),
      error_rate = 1 - mean(choice == 1)  # Assuming upper is correct
    )
  
  # Create diagnostic visualization
  diagnostic_values <- data.frame(
    Metric = c("Choice Bias", "Mean RT", "Min RT", "RT Skew", "Error Rate"),
    Value = c(diagnostics$prop_upper, diagnostics$mean_rt, 
              diagnostics$min_rt, diagnostics$rt_skew, diagnostics$error_rate),
    Expected = c(0.5, 0.7, 0.2, 0.5, 0.2),  # Example expected values
    Type = c("Proportion", "Time", "Time", "Ratio", "Proportion")
  )
  
  ggplot(diagnostic_values, aes(x = Metric, y = Value)) +
    geom_point(size = 5, color = "darkblue") +
    geom_point(aes(y = Expected), size = 3, color = "gray50", shape = 1) +
    geom_segment(aes(xend = Metric, y = Expected, yend = Value),
                 alpha = 0.5, linetype = "dashed") +
    facet_wrap(~ Type, scales = "free_y", nrow = 1) +
    labs(title = title,
         subtitle = "Blue dots = observed, gray circles = expected",
         y = "Value") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

# Example diagnostic
diagnostic_example <- create_diagnostic_plot(
  baseline_data,
  "Baseline Model Diagnostics"
)
print(diagnostic_example)
```

## 🎯 Key Takeaways

1.  **Parameters have distinct signatures**: Each parameter affects the data in characteristic ways

    -   Drift → Accuracy and mean RT

    -   Threshold → Speed-accuracy tradeoff

    -   Starting point → Choice bias and RT asymmetry

    -   Variability → Distribution shape and skew

2.  **Parameters interact**: The effect of one parameter depends on others

    -   Low drift amplifies bias effects

    -   High thresholds amplify drift effects

    -   Variability parameters modulate these relationships

3.  **Practical interpretation requires context**:

    -   Task demands (speed vs accuracy emphasis)

    -   Stimulus properties (difficulty, discriminability)

    -   Individual differences (age, expertise, clinical status)

4.  **Model validation is crucial**: Always check that parameters make sense

    -   Are values in reasonable ranges?

    -   Do manipulations affect parameters as expected?

    -   Are parameter correlations sensible?

## 📚 Further Reading

-   Ratcliff & McKoon (2008): The diffusion decision model: Theory and data

-   Voss et al. (2013): Diffusion models in experimental psychology

-   Forstmann et al. (2016): Sequential sampling models in cognitive neuroscience
